---
title: "Worked Example: msPCA on mtcars"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Worked Example: msPCA on mtcars}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This vignette shows the basic workflow of `msPCA` on the built-in `mtcars` dataset.
We compute two sparse principal components, inspect the solution, and compare the sparse result with dense PCA.

## Install and load

Install the package directly from CRAN.
```{r install, eval=FALSE}
install.packages("msPCA")
```
You can then load the package as usual.
```{r load}
library(msPCA)
```

## Fit two sparse PCs

We work with the correlation matrix of `mtcars` and ask for two 4-sparse principal components under the default orthogonality constraint.

```{r}
Sigma <- cor(datasets::mtcars)

set.seed(42)
res <- mspca(Sigma, r = 2, ks = c(4, 4), type = "Sigma", feasibilityConstraintType = 0, verbose = FALSE)
print_mspca(res, Sigma)
```

## Working from the raw data matrix

By default (`type = "Sigma"`) the first argument is a covariance/correlation
matrix; set `type = "X"` to pass the raw data matrix instead (rows are
observations, columns are variables). With `type = "X"`, `msPCA` applies the
algorithm to the data directly: each product `Sigma %*% beta` is computed as
`t(X) %*% (X %*% beta) / (n - 1)`, so the `p x p` matrix is never formed. This is
mathematically equivalent to the matrix-input call but much more scalable when
the number of variables `p` greatly exceeds the number of observations `n` (it
avoids the `O(np^2)` cost of building `Sigma` and turns each iteration's
matrix–vector product from `O(p^2)` into `O(np)`).

The preprocessing arguments control which matrix is implicitly used:

- `center` (default `TRUE`) subtracts the column means,
- `scale` (default `FALSE`) divides columns by their standard deviation; set it to
  `TRUE` to operate on the correlation matrix,
- `divisor` chooses the covariance normalization, `"n-1"` (default, matching
  `cov`/`cor`) or `"n"`.

With `scale = TRUE` and `divisor = "n-1"`, the raw-data call reproduces the
correlation-matrix result above.

```{r}
X <- as.matrix(datasets::mtcars)

set.seed(42)
res_X <- mspca(X, r = 2, ks = c(4, 4), type = "X", scale = TRUE, verbose = FALSE)

# type = "X" results carry their own variance summary, so print_mspca() needs no matrix.
print_mspca(res_X)

# Same objective as the correlation-matrix call:
c(matrix_input = res$objective_value, data_input = res_X$objective_value)
```

The same dual input is available for the single-component `tpm()`.

## Orthogonality versus zero correlation

Sparse PCA typically requires a constraint to avoid redundancy between the PCs. Traditionally, this is done by enforcing orthogonality of the loading vectors, which is the default in `mspca`. Another notion of non-redundancy is to enforce zero pairwise correlation between the PCs.
The package allows for both options, and the choice can lead to different solutions when the variables are strongly correlated.
`feasibilityConstraintType = 0` (default) enforces orthogonality of the loading vectors.
`feasibilityConstraintType = 1` instead enforces zero pairwise correlation between the resulting components.

```{r}
res_corr <- mspca(Sigma, r = 2, ks = c(4, 4), type = "Sigma", feasibilityConstraintType = 1, verbose = FALSE)
print_mspca(res_corr, Sigma)
```

## Diagnostics

The package provides helper functions for checking feasibility and summarizing variance explained.
Below, we report the same diagnostic checks for each fitted solution.

```{r}
cat("Diagnostics for res (feasibilityConstraintType = 0)\n")
feasibility_violation_off(Sigma, res$x_best, feasibilityConstraintType = 0)
feasibility_violation_off(Sigma, res$x_best, feasibilityConstraintType = 1)
fraction_variance_explained(Sigma, res$x_best)
fraction_variance_explained_perPC(Sigma, res$x_best)

cat("\nDiagnostics for res_corr (feasibilityConstraintType = 1)\n")
feasibility_violation_off(Sigma, res_corr$x_best, feasibilityConstraintType = 0)
feasibility_violation_off(Sigma, res_corr$x_best, feasibilityConstraintType = 1)
fraction_variance_explained(Sigma, res_corr$x_best)
fraction_variance_explained_perPC(Sigma, res_corr$x_best)
```

## Comparison with dense PCA

For reference, the first two dense principal components explain more variance, but they are not sparse.

```{r}
pca_res <- prcomp(datasets::mtcars, scale. = TRUE)
fraction_variance_explained(Sigma, pca_res$rotation[, 1:2])
```

## Interpretation

Sparse PCA typically trades some explained variance for a much more interpretable loading pattern.
For a quick summary of the fitted components, `print_mspca()` is usually the most useful first diagnostic.
