---
title: "Benchmarking against other sparse PCA packages"
output: rmarkdown::html_vignette
bibliography: ../thebib.bib
---

<!--
  %\VignetteIndexEntry{Benchmarking against other sparse PCA packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette gives the benchmarking protocol and results behind the summary in the package
description. It compares `msPCA` against eight competing implementations, drawn from seven
packages, on four datasets, and records how each competitor was configured. That configuration
matters more than usual here, because the packages parameterize sparsity in mutually incompatible
ways.

Reproducing the comparison requires eight additional R
packages, plus `callr` and a Python installation, and the memory measurements require macOS or Linux.
The scripts are in the package repository under
[`replication/benchmarking/`](https://github.com/jeanpauphilet/msPCA/tree/main/replication/benchmarking).

## Getting the data

All four datasets are obtainable from R without manual downloads or accounts, so the
comparison is reproducible end to end:

```{r}
## mtcars -- base R
X_mtcars     <- as.matrix(datasets::mtcars)
Sigma_mtcars <- cor(X_mtcars)

## Pitprops -- the 13 x 13 correlation matrix of Jeffers (1967), Table 1.
## It is transcribed literally at the top of
## replication/benchmarking/notebook_pitprops.R; source that file to obtain `pitprops`.
Sigma_pitprops <- pitprops
## Methods that require a data matrix instead use pseudo-data drawn from this
## correlation structure (see "How each method was configured" below).
set.seed(42)
X_pitprops <- MASS::mvrnorm(n = 500, mu = rep(0, ncol(pitprops)),
                            Sigma = pitprops)

## Breast cancer -- one-off download via PMA, then keep the 500 highest-
## variance genes, following the preprocessing of Witten et al. (2009).
X            <- t(PMA::download_breast_data(
  url = "https://tibshirani.su.domains/PMA/breastdata.rda")$rna)
X_breast     <- X[, order(apply(X, 2, var), decreasing = TRUE)[1:500]]
Sigma_breast <- cor(X_breast)          # n = 89, p = 500

## Riboflavin -- ships with the hdi package
data("riboflavin", package = "hdi")
X_ribo <- as.matrix(riboflavin$x)      # n = 71, p = 4088
```

## Methods compared

We compare `msPCA::mspca()` against eight functions drawn from seven competing packages. Each
entry below names the exact function called, not merely the package, since some packages expose
more than one sparse PCA routine.

| Function | Reference | Language | Approach |
|---|---|---|---|
| `elasticnet::spca()` | @zou2006sparse | R | elastic-net regularization, LARS-like |
| `PMA::SPC()` | @witten2009penalized | R | penalized matrix decomposition |
| `sparsepca::spca()` | @erichson2020sparse | R | variable projection with $\ell_1$ penalty |
| `mixOmics::spca()` | @rohart2017mixomics | R | regularized SVD [@shen2008sparse] |
| `nsprcomp::nsprcomp()` | @sigg2019nsprcomp | R | expectation-maximization [@sigg2008expectation] + deflation |
| `nsprcomp::nscumcomp()` | @sigg2019nsprcomp | R | cumulative variant, all components jointly |
| `amanpg::spca.amanpg()` | @chen2020proximal | R | alternating manifold proximal gradient |
| `sklearn.decomposition.SparsePCA()` | @pedregosa2011scikit | Python | dictionary learning [@mairal2009online] |

`scikit-learn` is called from R via the `reticulate` package [@ushey2023reticulate]. Together
these cover the main methodological approaches to sparse PCA: elastic-net regularization,
penalized matrix decomposition, variable-projection optimization, regularized SVD, deflation,
Riemannian proximal gradient, and dictionary learning.

### How each function obtains multiple components, and what its tuning parameter controls

The comparison turns on two things that are easy to conflate: how a method gets from one sparse
component to $r$ of them, and what the parameter the user actually turns is doing. The second
column below states the mechanism; the last two name the parameter we tune and classify it as
controlling **sparsity** or as an **orthogonality penalty**.

| Function | Multiple-component mechanism | Parameter tuned | What it controls |
|---|---|---|---|
| `msPCA::mspca()` | Iterative deflation heuristic; all $r$ components are refined together, with the non-redundancy constraint enforced across every pair and its penalty tightened over the iterations | `ks` | **Sparsity** --- exact cardinality, one budget per component |
| `elasticnet::spca()` | All $r$ jointly: alternates an orthonormal Procrustes update of an auxiliary factor with per-component elastic-net regressions. Orthonormality is imposed on the auxiliary factor, not on the returned sparse loadings | `para` (with `sparse = "varnum"`) | **Sparsity** --- exact cardinality per component |
| `PMA::SPC()` | Sequential rank-1 penalized matrix decompositions with deflation. `orth = TRUE` selects the variant of Section 3.2 of @witten2009penalized, which requires the *left* factors $\boldsymbol{u}_k$ to be mutually orthogonal --- not the sparse loadings $\boldsymbol{v}_k$ that the function returns and that we score | `sumabsv` | **Sparsity** --- an $\ell_1$ bound on each loading vector, not a cardinality |
| `sparsepca::spca()` | All $r$ jointly, by variable projection on the SPCA objective | `alpha` | **Sparsity** --- magnitude of the $\ell_1$ penalty, with no per-component control |
| `mixOmics::spca()` | Sequential regularized rank-1 SVDs with deflation | `keepX` | **Sparsity** --- exact number of variables retained per component |
| `nsprcomp::nsprcomp()` | Sequential cardinality-constrained components via an EM-style iteration, combined with the deflation of @mackey2008deflation | `k` | **Sparsity** --- exact cardinality per component |
| `nsprcomp::nscumcomp()` | All $r$ jointly, under one cumulative budget | `k`, `gamma` | `k`: **sparsity**, as a *total* budget $\sum_t k_t$; `gamma`: **orthogonality penalty**, which we tune (see below) rather than leaving at its default of 0 |
| `amanpg::spca.amanpg()` | All $r$ jointly: alternates a proximal-gradient step on the sparse loadings with a manifold update of an orthonormal factor on the Stiefel manifold | `lambda1` | **Sparsity** --- a vector of per-component $\ell_1$ penalties |
| `sklearn...SparsePCA()` | All `n_components` jointly, by dictionary learning | `alpha` | **Sparsity** --- magnitude of the $\ell_1$ penalty on the loadings |

Compared with `msPCA::mspca()`:

**Among the competitors, only `nsprcomp::nscumcomp()` offers any control over non-redundancy
(orthogonality)**, through the soft penalty `gamma`, whose default of 0 applies no penalty at all.
For a fair comparison, we tune `gamma` along a grid: among the values that bring the orthogonality
violation within 1e-4 (our default tolerance in `msPCA`), we keep the one with the highest FVE.

**Not all methods control sparsity in the same way.** Four functions --- `msPCA::mspca()`,
`elasticnet::spca()`, `mixOmics::spca()` and `nsprcomp::nsprcomp()` --- accept an exact cardinality
per component. The others expose either a penalty parameter that controls sparsity only indirectly
(`sparsepca::spca()`, `amanpg::spca.amanpg()`, `sklearn...SparsePCA()`), an $\ell_1$ bound
(`PMA::SPC()`), or a single cumulative sparsity budget covering all components together
(`nsprcomp::nscumcomp()`). For these methods, we tune the hyperparameter so as to (i) return the
requested number of non-empty components, $r$, and (ii) come as close as possible to the requested
total sparsity, $\sum_t k_t$.

## Environment

Experiments were conducted on a MacBook Air with an Apple M2 chip and 24 GB of RAM, with
R version 4.6.0 (2026-04-24), `msPCA` v0.5.1, `elasticnet` v1.3, `PMA` v1.2.4, `sparsepca`
v0.1.2, `mixOmics` v6.36.0, `nsprcomp` v0.5.1.2, `amanpg` v0.3.4, `reticulate` v1.46.0,
`callr` v3.8.0, and `scikit-learn` v1.6.1 (Python 3.11). Full session information is recorded in
[`replication/sessionInfo.txt`](https://github.com/jeanpauphilet/msPCA/blob/main/replication/sessionInfo.txt).

## Configuration

The table above gives each function's mechanism and tuning parameter. What remains to be fixed
is which input each one receives, since some accept a covariance or correlation matrix, some
require the data matrix, and some take either.

- **`msPCA::mspca()`** --- run with both `type = "Sigma"` and `type = "X"` on all four datasets.
  On riboflavin ($n \ll p$) the $p \times p$ covariance matrix is rank-deficient, making
  `type = "X"` the natural choice; we report both modes to quantify the runtime and memory gain.
- **`elasticnet::spca()`** --- takes a list of cardinality budgets (one per PC) via
  `sparse = "varnum"`, and applies interchangeably to $\boldsymbol{\Sigma}$ (`type = "Gram"`) or
  $\boldsymbol{X}$ (`type = "predictor"`). We report both input modes on mtcars, Pitprops and breast
  cancer.
- **`PMA::SPC()`** --- applied to the **data matrix** on mtcars, Pitprops and breast cancer, as
  documented (`x` is "a data matrix of dimension $n \times p$", and `SPC()` centers its columns itself). On
  Pitprops that means the pseudo-data, since no real data matrix exists. `sumabsv` bounds the
  $\ell_1$ norm of each loading vector (subject to unit $\ell_2$ norm) and must lie between 1
  and $\sqrt{p}$; we tune it per dataset to bring per-component sparsity as close to the target
  as the parameterization allows. We set `orth = TRUE`, which selects the multiple-component
  variant of Section 3.2 of @witten2009penalized.
- **`sparsepca::spca()`** --- applied to $\boldsymbol{\Sigma}$. It exposes only the magnitude of the
  $\ell_1$ penalty `alpha`, which we tune per dataset to achieve a total sparsity
  $\sum_t k_t$ close to the target. This method tends to return PCs with very unequal sparsity
  levels, the first typically denser than the rest.
- **`mixOmics::spca()`** --- takes the data matrix and the list of per-component budgets
  `keepX`, so it hits the cardinality target exactly.
- **`nsprcomp::nsprcomp()`** and **`nsprcomp::nscumcomp()`** --- two functions from the same
  package, both based on the EM algorithm of @sigg2008expectation. 
  `nsprcomp()` computes the leading sparse PC then
  applies the deflation procedure of @mackey2008deflation for subsequent PCs; it takes the data
  matrix and a vector of per-component cardinalities. `nscumcomp()` solves for all $r$ PCs
  simultaneously under a single cumulative budget $\sum_t k_t$, which it distributes as it sees
  fit, plus the penalty `gamma` on divergence from orthonormality.
- **`amanpg::spca.amanpg()`** --- accepts either input and is told which via `type`: `type = 0`
  declares `z` to be a data matrix, any non-zero value declares it a covariance matrix. We use
  `type = 1` with the correlation matrix on mtcars and Pitprops, where the correlation matrix is
  the natural object, and `type = 0` with the data matrix on breast cancer. It computes all $r$
  components jointly, alternating a proximal-gradient step on the sparse loadings with a manifold
  update of an orthonormal factor on the Stiefel manifold; `lambda1` is a vector of per-component
  $\ell_1$ penalties, which we tune per dataset, and we set `lambda2 = Inf`.
- **`sklearn.decomposition.SparsePCA()`** --- called via `reticulate`. The class computes
  `n_components` components simultaneously using dictionary learning with an $\ell_1$ penalty
  `alpha` on the loadings, applied to the data matrix. We tune `alpha` per dataset and fix
  `random_state` to the repetition's seed.

For Pitprops only the correlation matrix is available, so we generate pseudo-data via
`MASS::mvrnorm()` with the Pitprops matrix as the population covariance ($n = 500$) for the
methods that require $\boldsymbol{X}$. FVE and the orthogonality
violation are always evaluated against the published Pitprops correlation matrix, never against
the covariance of the pseudo-data.

Every tuning parameter is selected by the scripts rather than fixed by hand: each is swept over a
grid and the value whose realized sparsity comes closest to the target is kept, with ties broken
on FVE and candidates that leave a component empty discarded. The selected values appear beneath
each table. Tuning is excluded from all reported runtimes and memory figures.

### How runtime and memory are measured

Every repetition of every method runs in its **own fresh R process**, driven by
[`replication/benchmarking/bench_utils.R`](https://github.com/jeanpauphilet/msPCA/blob/main/replication/benchmarking/bench_utils.R).
Isolation is necessary because peak memory is read as a process-level high-water mark: several
methods in one session, or even several repetitions of one method, contaminate each other.

Memory is obtained from `getrusage(RUSAGE_SELF).ru_maxrss`. R-level
metrics such as `gc()` or `bench::mark(mem_alloc)` are deliberately **not** used: `msPCA` does
its work in `RcppEigen` buffers and `scikit-learn` in the embedded CPython heap, neither of
which R's garbage collector observes, so an R-level metric would under-report memory usage for those two
methods. The **Memory** column below reports the
working set, defined as the size of the matrix handed to the method plus the additional peak
RSS the method allocates on top of it. 

Each method is repeated five times with a different seed. All four measured quantities are
reported as medians. For FVE, runtime, and memory, we also report a bracketed min--max range wherever the spread
is visible at the printed precision; for the orthogonality violation, we report the worst (i.e., largest) value.

```{r}
# Representative call pattern; see replication/benchmarking/ for the full scripts.
set.seed(42)
res <- mspca(Sigma, r = 3, ks = rep(4, 3), verbose = FALSE)

fraction_variance_explained(Sigma, res$x_best)
feasibility_violation_off(Sigma, res$x_best, 0)
res$runtime
```

## Results

### mtcars

`datasets::mtcars` from base R ($p = 11$ variables, $n = 32$ observations) serves as a simple
reproducible baseline. We extract $r = 3$ sparse PCs targeting $k = 4$ per component.

| Method | $(k_1,k_2,k_3)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (4,4,4) | 0.829 | 9.4e-05 | 0.189 | 0.4 [0.4, 0.7] |
| `msPCA::mspca` $(\boldsymbol{X})$ | (4,4,4) | 0.829 | 9.4e-05 | 0.245 | 0.2 |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (4,4,4) | 0.753 | 0.117 | 0.093 | 21 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (4,4,4) | 0.754 | 0.117 | 0.112 | 23 [23, 28] |
| `PMA::SPC` | (4,4,5) | 0.551 | 0.221 | 0.009 | 0.6 [0.6, 3.6] |
| `sparsepca::spca` | (6,4,1) | 0.581 | 0.133 | 0.032 [0.031, 0.036] | 17 |
| `mixOmics::spca` | (4,4,4) | 0.633 | 0.133 | 0.003 | 0.5 |
| `nsprcomp::nsprcomp` | (4,4,4) | 0.832 | 0.078 | 0.004 [0.003, 0.005] | 0.1 |
| `nsprcomp::nscumcomp` | (5,5,2) | 0.824 [0.798, 0.825] | 1.1e-05 (max 3.3e-05) | 0.241 | 30 |
| `amanpg::spca.amanpg` | (4,4,1) | 0.544 | 0.047 | 0.007 | 1.5 |
| `sklearn...SparsePCA` | (4,3,4) | 0.827 | < 1e-15 | 0.012 [0.010, 0.015] | 0.0 [0.0, 0.1] |
| PCA (dense) | (11,11,11) | 0.899 | < 1e-15 | 0.001 | 0.5 |

Tuning: `sumabsv = 1.85`, `alpha = 0.0045` (`sparsepca`), `lambda1 = c(4.5, 5.4, 0.1)`,
`alpha = 3.2` (`sklearn`), `gamma = 1e6`.

`nscumcomp` does well here: it is within 0.01 FVE of the best sparse method and comfortably inside
the feasibility tolerance. However, it returns a sparsity pattern of (5,5,2) rather than the
requested (4,4,4), and it is the only method whose FVE varies across seeds, spanning 0.798--0.825.

Because $p = 11$, every method's numerical working set is a few kilobytes, so the memory column measures
fixed overheads: most methods sit at or below 1.5 MB, while `elasticnet`, `sparsepca` and `nscumcomp`
carry a floor of 17--30 MB that is already present at this size.

### Pitprops

The Pitprops dataset [@jeffers1967two] ($p = 13$ physical measurements on $n = 180$ timber
specimens) is a classic sparse PCA benchmark [e.g., @zou2006sparse]. We extract $r = 6$ sparse
PCs targeting $k = 4$ nonzero loadings per component, as in @zou2006sparse.

| Method | $(k_1, \ldots, k_6)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (4,4,4,4,4,4) | 0.791 [0.786, 0.822] | 9.7e-05 (max 9.8e-05) | 0.647 | 0.7 [0.4, 0.8] |
| `msPCA::mspca` $(\boldsymbol{X})$ | (4,4,4,4,4,4) | 0.794 [0.784, 0.801] | 9.4e-05 (max 9.9e-05) | 3.958 | 0.3 [0.3, 0.9] |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (4,4,4,4,4,4) | 0.792 | 0.241 | 0.070 | 21 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (4,4,4,4,4,4) | 0.790 | 0.322 | 0.158 | 73 [70, 83] |
| `PMA::SPC` | (3,4,4,4,3,4) | 0.714 | 0.491 | 0.021 | 15 |
| `sparsepca::spca` | (7,3,4,2,4,4) | 0.756 | 0.353 | 0.039 | 28 |
| `mixOmics::spca` | (4,4,4,4,4,4) | 0.788 | 0.215 | 0.010 [0.009, 0.011] | 0.6 [0.5, 0.6] |
| `nsprcomp::nsprcomp` | (4,4,4,4,4,4) | 0.816 [0.813, 0.833] | 0.563 (max 1.105) | 0.025 [0.023, 0.028] | 34 [28, 42] |
| `nsprcomp::nscumcomp` | (5,6,5,2,4,2) | 0.819 [0.814, 0.830] | 5.1e-05 (max 1.0e-04) | 1.098 | 68 [64, 129] |
| `amanpg::spca.amanpg` | (3,3,4,3,4,6) | 0.863 | 1.654 | 0.011 [0.011, 0.013] | 4.5 |
| `sklearn...SparsePCA` | (7,3,4,3,5,4) | 0.907 | 0.815 | 0.061 | 0.1 |
| PCA (dense) | (13, ..., 13) | 0.877 | < 1e-15 | 0.001 | 0.4 |

Tuning: `sumabsv = 1.55`, `alpha = 0.005` (`sparsepca`),
`lambda1 = c(4.80, 4.40, 1.95, 5.50, 0.20, 0.05)`, `alpha = 2.6` (`sklearn`), `gamma = 1e7`.

This dataset is the one where the feasibility differences are starkest. With $p = 13$ and 24
nonzero coordinates in total, disjoint-support solutions do not exist, so methods that do not
explicitly enforce orthogonality --- the deflation-based ones in particular --- struggle:
at the requested cardinality `elasticnet`, `mixOmics` and `nsprcomp` return violations of
0.215--0.563 against 9.7e-05 for `msPCA`. `nscumcomp` also reaches the tolerance (with
$\gamma = 10^7$ against a default of 0), but at a fortyfold slowdown relative to `nsprcomp`
(1.098 s against 0.025 s) and without achieving the requested sparsity.

Pitprops is also the one case where `type = "X"` is the wrong choice: with $n = 500$
pseudo-observations and $p = 13$ the data matrix is larger than the correlation matrix, and
runtime is roughly six times worse. Across seeds `msPCA` ranges over 0.786--0.822 in FVE, the
widest spread it shows on any of the four datasets; the orthogonality violation stays below 1e-04
throughout.

### Breast cancer gene expression

Gene expression data from @chin2006genomic ($n = 89$ tumor samples, 19,672 genes). Following
the preprocessing of @witten2009penalized we retain the $p = 500$ genes with the highest
marginal variance, yielding a high-dimensional ($p > n$) benchmark. We set $r = 3$ and target
$k = 20$ nonzero loadings per component.

| Method | $(k_1, k_2, k_3)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (20,20,20) | 0.093 [0.088, 0.095] | 7.5e-05 (max 9.8e-05) | 27.244 [14.119, 27.602] | 9.4 [6.6, 11.3] |
| `msPCA::mspca` $(\boldsymbol{X})$ | (20,20,20) | 0.095 [0.088, 0.095] | 7.5e-05 (max 9.8e-05) | 13.268 [6.790, 13.399] | 3.2 [2.8, 4.4] |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (20,20,20) | 0.043 | 0.040 | 16.104 | 192 [149, 220] |
| `elasticnet::spca` $(\boldsymbol{X})$ | (20,20,20) | 0.043 | 0.021 | 0.816 | 79 [79, 127] |
| `PMA::SPC` | (20,22,13) | 0.042 | 0.008 | 0.041 | 33 |
| `sparsepca::spca` | (43,14,22) | 0.082 | < 1e-15 | 4.841 | 142 [46, 176] |
| `mixOmics::spca` | (20,20,20) | 0.077 | < 1e-15 | 0.032 | 7.0 [6.9, 27.6] |
| `nsprcomp::nsprcomp` | (20,20,20) | 0.083 [0.070, 0.084] | < 1e-15 | 0.032 [0.030, 0.035] | 48 |
| `nsprcomp::nscumcomp` | (29,23,8) | 0.081 [0.080, 0.085] | < 1e-15 | 0.229 [0.219, 0.261] | 55 |
| `amanpg::spca.amanpg` | (68,1,2) | 0.045 | < 1e-15 | 0.071 | 50 [49, 65] |
| `sklearn...SparsePCA` | (22,3,1) | 0.020 [0.020, 0.032] | < 1e-15 | 0.079 | 0.6 [0.5, 1.3] |
| PCA (dense) | (500,500,500) | 0.352 | < 1e-15 | 0.010 | 2.6 [2.5, 3.6] |

Tuning: `sumabsv = 3.3`, `alpha = 0.0035` (`sparsepca`), `lambda1 = c(4.5, 1.8, 3.3)`,
`alpha = 15` (`sklearn`), `gamma = 316`.

This is where the penalty-parameterized methods break down most clearly in terms of achieved
sparsity. Against a request of (20,20,20), `amanpg` returns (68,1,2), `sklearn` (22,3,1) and
`sparsepca` (43,14,22); no value on their grids does better. Only the four functions taking a
per-component cardinality budget return the desired sparsity levels. Feasibility, by contrast, is
uninformative here: with 60 nonzeros over $p = 500$ disjoint supports are easy to find, and most
methods return solutions that are trivially orthogonal. `msPCA` is the exception, at 7.5e-05: it
stops once the violation is inside its 1e-4 tolerance rather than driving it to zero.

This is the smallest dataset at which the memory column measures the algorithms rather than fixed
overheads, and `msPCA` on the data matrix is the lightest of the methods that meet the budget:
3.2 MB, against 7.0 MB for `mixOmics`, 9.4 MB for `msPCA` on the correlation matrix, 48 MB for
`nsprcomp`, and 79 and 192 MB for `elasticnet`. (`sparsepca` at 142 MB and `nscumcomp` at 55 MB do
not meet the budget.) The `type = "X"` saving is modest in absolute terms because
$\boldsymbol{\Sigma}$ is only about 2 MB at this size.

### Riboflavin

The riboflavin (vitamin B2) production dataset [@buhlmann2014high] comprises log-transformed
expression levels of $p = 4{,}088$ genes in $n = 71$ samples of *Bacillus subtilis*. This is a
challenging $p \gg n$ benchmark: the empirical correlation matrix is rank-deficient
(rank $\leq 70$). We set $r = 2$ and target $k = 20$ nonzero loadings per component. To avoid
time-consuming parameter tuning, we consider only methods that accept exact cardinality
budgets.

| Method | $(k_1, k_2)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (20,20) | 0.008 [0.008, 0.009] | < 1e-15 | 57.226 [41.791, 87.700] | 640 |
| `msPCA::mspca` $(\boldsymbol{X})$ | (20,20) | 0.008 [0.008, 0.009] | < 1e-15 | 1.557 [0.777, 2.177] | 44 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (20,20) | 0.004 | < 1e-15 | 6.303 | 512 [421, 577] |
| `mixOmics::spca` | (20,20) | 0.006 | < 1e-15 | 0.335 | 24 [24, 50] |
| `nsprcomp::nsprcomp` | (20,20) | 0.009 | < 1e-15 | 0.370 [0.265, 0.391] | 61 [57, 124] |
| PCA (dense) | (4088,4088) | 0.478 | < 1e-15 | 0.046 | 39 |

`msPCA` is run with `maxIter = 100` here, and every method other than `msPCA` $(\boldsymbol{\Sigma})$
takes the data matrix.

This dataset isolates the benefit of `type = "X"`. Passing the raw data matrix replaces the
$O(p^2)$ product $\boldsymbol{\Sigma}\boldsymbol{\beta}$ with the $O(np)$ two-pass product
$\boldsymbol{X}^\top(\boldsymbol{X}\boldsymbol{\beta})/(n-1)$ and never forms the covariance matrix: the working set
falls from 640 MB to 44 MB and median runtime from 57.2 seconds to 1.6 seconds, at identical FVE and
feasibility. The correlation matrix accounts for 128 MB of the difference, the rest being the
`Eigen` copy of it and the dense intermediates the $O(p^2)$ path requires. Both rows vary by a
factor of two or more in runtime across seeds, so those ratios are good to one significant figure
at most.

The memory figures are far steadier: `msPCA` $(\boldsymbol{\Sigma})$ spans 639--641 MB over the same
repetitions, whereas `elasticnet`'s five repetitions span 421--577 MB and its median moves between
runs of the identical script. `msPCA` allocates in `Eigen`, where the footprint follows from the
algorithm; the pure-R methods allocate on R's heap, where the peak depends on garbage-collector
timing. Together with the Pitprops result the rule is just the relative size of the two inputs:
`type = "X"` when $n \ll p$, `type = "Sigma"` when $p \lesssim n$.

## Discussion

**Sparsity control.** `PMA::SPC()`, `sparsepca::spca()`, `nsprcomp::nscumcomp()`,
`amanpg::spca.amanpg()` and `sklearn.decomposition.SparsePCA()` give no direct control over the
sparsity of each loading vector. `PMA::SPC()` exposes only a global $\ell_1$ bound and `sklearn`
only a global $\ell_1$ penalty;
`sparsepca::spca()`, `nscumcomp()` and `amanpg` tend to return unbalanced components, with one PC
carrying most of the budget. None can be calibrated to an exact per-component target, which is
what makes an apples-to-apples comparison awkward. Where the discussion below needs methods that
meet the budget exactly, those are `msPCA::mspca()`, `elasticnet::spca()`, `mixOmics::spca()` and
`nsprcomp::nsprcomp()`.

**Feasibility.** Two things have to be separated here, because the answer depends on whether
disjoint supports exist at all. On breast cancer and riboflavin, where $p$ is large relative to
the total budget, orthogonality is nearly free: most methods return exactly orthogonal loadings
simply because their supports do not overlap. On mtcars ($p = 11$, 12 nonzeros) and Pitprops
($p = 13$, 24 nonzeros) no disjoint solution exists, and that is where the differences appear.

There, `msPCA::mspca()` reaches 1e-4 while returning the requested cardinality exactly, and does
so at every repetition: the largest violation it produced anywhere in the benchmark is 9.9e-05.
Two other methods reach the tolerance, both with qualifications.
`nsprcomp::nscumcomp()` reaches it on both datasets, but only with `gamma` driven to $10^6$--$10^7$, returning
(5,5,2) and (5,6,5,2,4,2) against budgets of (4,4,4) and (4,4,4,4,4,4), and its worst Pitprops
repetition is 1.03e-04. `sklearn...SparsePCA()` reaches it on mtcars by returning 11 nonzeros in
total rather than the requested 12, which at $p = 11$ admits a disjoint solution and so makes
orthogonality free. Every other method, at the requested cardinality, returns violations between
0.08 and 0.56 on these two datasets.

The claim supported by this is narrower than "only `msPCA` can be feasible", and more useful: on
the two datasets where feasibility actually binds, it is the only method that delivers feasibility
and the requested sparsity together, without a search over a penalty parameter to get there.

**Variance explained.** Among the methods that meet the requested cardinality,
`msPCA::mspca()` and `nsprcomp::nsprcomp()` are generally the best performers, followed by
`mixOmics::spca()` and `elasticnet::spca()`. The FVE cost of enforcing feasibility therefore looks
small: the largest gap between `msPCA` and the best budget-respecting competitor is about 0.025
FVE, on Pitprops, where `nsprcomp` reaches 0.816 at an orthogonality violation of 0.563.

Several methods post higher FVE while missing the budget, and those figures should be read in
that light: `sklearn...SparsePCA()` reaches 0.907 on Pitprops at (7,3,4,3,5,4) with a violation of
0.815, and `amanpg::spca.amanpg()` 0.863 at a violation of 1.654. The exception is `nscumcomp` at
0.824 on mtcars with (5,5,2), which is both high and feasible, and is the most substantive
competition in the comparison.

**Reproducibility.** Repeating each method under several seeds separates the implementations that
returned identical results at every seed (`elasticnet`, `PMA`, `sparsepca`, `mixOmics`, `amanpg`,
dense PCA) from the rest. `nsprcomp` and `nscumcomp` use random initialization and move between
runs; `nsprcomp`'s orthogonality violation on Pitprops spans 0.56--1.10.
`sklearn...SparsePCA()`, seeded from the repetition's seed, also moves (0.020--0.032 FVE on breast
cancer). `msPCA` uses random restarts and is likewise not bit-reproducible across seeds, but its
variation is confined to the objective: the feasibility target was met at every seed on every
dataset, the largest violation observed anywhere being 9.9e-05.

**Memory.** On the two high-dimensional datasets, where the memory column reflects the
algorithms rather than fixed overheads, `msPCA` is among the lightest of the sparse methods that
respect an exact cardinality budget. On breast cancer it uses 3.2 MB on the data matrix against
79 MB for `elasticnet` on the same input and 192 MB in Gram mode, 7.0 MB for `mixOmics` and 48 MB
for `nsprcomp` (with `sparsepca` at 142 MB and `nscumcomp` at 55 MB, neither meeting the budget).
On riboflavin it uses 44 MB against 512 MB for `elasticnet` and 61 MB for `nsprcomp`, though
`mixOmics` is lighter still at 24 MB. Within `msPCA`, the
$\boldsymbol{\Sigma}$-versus-$\boldsymbol{X}$ choice matters more than the choice of package: the
same function spans 640 MB to 44 MB on riboflavin depending on the interface.

**Scalability.** `msPCA::mspca()` scales to datasets with $p$ in the thousands in reasonable
time, and the `type = "X"` interface yields substantial runtime and memory gains when
$n \ll p$: roughly 35x faster and 15x lighter on riboflavin. When $n > p$, as in the Pitprops
pseudo-data, the ordering reverses and `type = "Sigma"` is preferable.

## References
