---
title: "Benchmarking against other sparse PCA packages"
output: rmarkdown::html_vignette
bibliography: thebib.bib
---

<!--
  %\VignetteIndexEntry{Benchmarking against other sparse PCA packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette gives the benchmarking protocol and results behind the summary in the package
description. It compares `msPCA` against seven competing implementations on four real datasets,
and records how each competitor was configured. That configuration matters more than usual here,
because the packages parameterise sparsity in mutually incompatible ways.

Chunks are not evaluated at build time. Reproducing the comparison requires eight additional R
packages, `callr`, and a Python installation, and the memory measurements require macOS or Linux.
The scripts are in the package repository under
[`replication/benchmarking/`](https://github.com/jeanpauphilet/msPCA/tree/main/replication/benchmarking).

## Getting the data

All four datasets are obtainable from R without manual downloads or accounts, so the
comparison is reproducible end to end:

```{r}
## mtcars -- base R
X_mtcars     <- as.matrix(datasets::mtcars)
Sigma_mtcars <- cor(X_mtcars)

## Pitprops -- the 13 x 13 correlation matrix of Jeffers (1967), Table 1.
## It is transcribed literally at the top of
## replication/benchmarking/notebook_pitprops.R; source that file to obtain `pitprops`.
Sigma_pitprops <- pitprops
## Methods that require a data matrix instead use pseudo-data drawn from this
## correlation structure (see "How each method was configured" below).
set.seed(42)
X_pitprops <- MASS::mvrnorm(n = 500, mu = rep(0, ncol(pitprops)),
                            Sigma = pitprops)

## Breast cancer -- one-off download via PMA, then keep the 500 highest-
## variance genes, following the preprocessing of Witten et al. (2009).
X            <- t(PMA::download_breast_data(
  url = "https://tibshirani.su.domains/PMA/breastdata.rda")$rna)
X_breast     <- X[, order(apply(X, 2, var), decreasing = TRUE)[1:500]]
Sigma_breast <- cor(X_breast)          # n = 89, p = 500

## Riboflavin -- ships with the hdi package
data("riboflavin", package = "hdi")
X_ribo <- as.matrix(riboflavin$x)      # n = 71, p = 4088
```

Only the breast cancer data touches the network, and `PMA::download_breast_data()` handles it
in a single call. Riboflavin ships inside `hdi`, `mtcars` is in base R, and the Pitprops
correlation matrix is transcribed in the benchmarking script, so no manual downloads or
accounts are needed anywhere.

## Methods compared

We compare `msPCA::mspca()` against seven competing functions. Each entry below names the exact
function called, not merely the package, since several packages expose more than one sparse PCA
routine and `nsprcomp` in particular provides two with quite different behaviour.

| Function | Reference | Language | Approach |
|---|---|---|---|
| `elasticnet::spca()` | @zou2006sparse | R | elastic-net regularization, LARS-like |
| `PMA::SPC()` | @witten2009penalized | R | penalized matrix decomposition |
| `sparsepca::spca()` | @erichson2020sparse | R | variable projection with $\ell_1$ penalty |
| `mixOmics::spca()` | @rohart2017mixomics | R | regularized SVD [@shen2008sparse] |
| `nsprcomp::nsprcomp()` | @sigg2019nsprcomp | R | expectation-maximization [@sigg2008expectation] + deflation |
| `nsprcomp::nscumcomp()` | @sigg2019nsprcomp | R | cumulative variant, all components jointly |
| `amanpg::spca.amanpg()` | @chen2020proximal | R | alternating manifold proximal gradient |
| `sklearn.decomposition.SparsePCA()` | @pedregosa2011scikit | Python | dictionary learning |

`scikit-learn` is called from R via the `reticulate` package [@ushey2023reticulate]. Together
these cover the main methodological approaches to sparse PCA: elastic-net regularization,
penalized matrix decomposition, variable-projection optimization, regularized SVD, deflation,
Riemannian proximal gradient, and dictionary learning.

### How each function obtains multiple components, and what its tuning parameter controls

The comparison turns on two things that are easy to conflate: how a method gets from one sparse
component to $r$ of them, and what the parameter the user actually turns is doing. The second
column below states the mechanism; the last two classify the parameter we tune as controlling
**sparsity**, **optimization convergence**, an **orthogonality penalty**, or a **feasibility
tolerance**.

| Function | Multiple-component mechanism | Parameter tuned | What it controls |
|---|---|---|---|
| `msPCA::mspca()` | Iterative deflation heuristic; all $r$ components are refined together, with the non-redundancy constraint enforced across every pair and its penalty tightened over the iterations | `ks` | **Sparsity** --- exact cardinality, one budget per component |
| `elasticnet::spca()` | All $r$ jointly: alternates an orthonormal Procrustes update of an auxiliary factor with per-component elastic-net regressions. Orthonormality is imposed on the auxiliary factor, not on the returned sparse loadings | `para` (with `sparse = "varnum"`) | **Sparsity** --- exact cardinality per component |
| `PMA::SPC()` | Sequential rank-1 penalized matrix decompositions with deflation. `orth = TRUE` selects the variant of Section 3.2 of @witten2009penalized, which requires the *left* factors $\boldsymbol{u}_k$ to be mutually orthogonal --- not the sparse loadings $\boldsymbol{v}_k$ that the function returns and that we score | `sumabsv` | **Sparsity** --- an $\ell_1$ bound on each loading vector, not a cardinality |
| `sparsepca::spca()` | All $r$ jointly, by variable projection on the SPCA objective | `alpha` | **Sparsity** --- magnitude of the $\ell_1$ penalty, with no per-component control |
| `mixOmics::spca()` | Sequential regularized rank-1 SVDs with deflation | `keepX` | **Sparsity** --- exact number of variables retained per component |
| `nsprcomp::nsprcomp()` | Sequential cardinality-constrained components via an EM-style iteration, combined with the deflation of @mackey2008deflation | `k` | **Sparsity** --- exact cardinality per component |
| `nsprcomp::nscumcomp()` | All $r$ jointly, under one cumulative budget | `k`, `gamma` | `k`: **sparsity**, as a *total* budget $\sum_t k_t$; `gamma`: **orthogonality penalty**, which we tune (see below) rather than leaving at its default of 0 |
| `amanpg::spca.amanpg()` | All $r$ jointly: alternates a proximal-gradient step on the sparse loadings with a manifold update of an orthonormal factor on the Stiefel manifold | `lambda1` | **Sparsity** --- a vector of per-component $\ell_1$ penalties |
| `sklearn...SparsePCA()` | All `n_components` jointly, by dictionary learning | `alpha` | **Sparsity** --- magnitude of the $\ell_1$ penalty on the loadings |

Two things stand out from that table.

**Every competing function is tuned through sparsity alone.** Four of them
(`elasticnet::spca()`, `mixOmics::spca()`, `nsprcomp::nsprcomp()` and `msPCA::mspca()`) accept an
exact cardinality per component and hit it. The others expose a penalty magnitude
(`sparsepca::spca()`, `amanpg::spca.amanpg()`, `sklearn...SparsePCA()`), an $\ell_1$ bound
(`PMA::SPC()`), or a single cumulative budget they distribute as they see fit
(`nsprcomp::nscumcomp()`). Their realised sparsity patterns consequently miss the target, often
by a wide margin and unevenly across components: (68,1,2) and (22,3,1) on breast cancer, against
a request of (20,20,20).

**Only `nsprcomp::nscumcomp()` offers any control over non-redundancy**, through the soft penalty
`gamma`, whose default of 0 applies no penalty at all. We tune `gamma` by a different rule from
the sparsity parameters: among values bringing the orthogonality violation within 1e-4, keep the
one with the highest FVE. The question that answers is "at equal feasibility, how much variance
does each method explain?"

`gamma` does work, on all three datasets where `nscumcomp` appears, but the penalties required are
extreme (10^6 on mtcars, 10^7 on Pitprops), the runtime cost reaches fortyfold, and the resulting
solutions still miss the requested per-component cardinality.

`PMA::SPC()`'s `orth` switch is not a second exception. It orthogonalises the left factors
$\boldsymbol{u}_k$; the sparse loadings $\boldsymbol{v}_k$ it returns carry no orthogonality guarantee either
way. No competing function exposes a feasibility *tolerance*: a value the user sets to say how
orthogonal the components must be, which the method then attains. That is what
`msPCA::mspca()`'s `feasibilityTolerance` (default 1e-4) is, and `nscumcomp` reaches the same
target only through a search over `gamma` conducted outside the package.

Convergence controls (`maxIter` and `stallingTolerance` in `msPCA::mspca()`,
`em_tol`/`em_maxiter`/`nrestart` in `nsprcomp`, `tol`/`maxiter` in `amanpg`, `tol`/`max_iter` in
`scikit-learn`) govern **optimization convergence** rather than sparsity or feasibility. All are
left at package defaults, except `maxIter = 100` for `msPCA::mspca()` on riboflavin.

## Environment

Experiments were conducted on a MacBook Air with an Apple M2 chip and 24 GB of RAM, with
R version 4.6.0 (2026-04-24), `msPCA` v0.5.1, `elasticnet` v1.3, `PMA` v1.2.4, `sparsepca`
v0.1.2, `mixOmics` v6.36.0, `nsprcomp` v0.5.1.2, `amanpg` v0.3.4, `reticulate` v1.46.0,
`callr` v3.8.0, and `scikit-learn` v1.6.1 (Python 3.11). Full session information is recorded in
[`replication/sessionInfo.txt`](https://github.com/jeanpauphilet/msPCA/blob/main/replication/sessionInfo.txt).

## Configuration

The table above gives each function's mechanism and tuning parameter. What remains to be fixed
is which input each one receives, since some accept a covariance or correlation matrix, some
require the data matrix, and some take either.

- **`msPCA::mspca()`** --- run with both `type = "Sigma"` and `type = "X"` on all four datasets.
  On riboflavin ($n \ll p$) the $p \times p$ covariance matrix is rank-deficient, making
  `type = "X"` the natural choice; we report both modes to quantify the runtime and memory gain.
- **`elasticnet::spca()`** --- takes a list of cardinality budgets (one per PC) via
  `sparse = "varnum"`, and applies interchangeably to $\boldsymbol{\Sigma}$ (`type = "Gram"`) or
  $\boldsymbol{X}$ (`type = "predictor"`). We report both input modes on mtcars, Pitprops and breast
  cancer.
- **`PMA::SPC()`** --- applied to the **data matrix** on all three datasets, as documented
  (`x` is "a data matrix of dimension $n \times p$", and `SPC()` centres its columns itself). On
  Pitprops that means the pseudo-data, since no real data matrix exists. `sumabsv` bounds the
  $\ell_1$ norm of each loading vector (subject to unit $\ell_2$ norm) and must lie between 1
  and $\sqrt{p}$; we tune it per dataset to bring per-component sparsity as close to the target
  as the parameterization allows. We set `orth = TRUE`, which selects the multiple-component
  variant of Section 3.2 of @witten2009penalized; note that this orthogonalises the left factors
  $\boldsymbol{u}_k$ rather than the sparse loadings $\boldsymbol{v}_k$ we extract, so the
  orthogonality violations reported for `PMA::SPC()` below are not a failure to deliver something
  the function promised.
- **`sparsepca::spca()`** --- applied to $\boldsymbol{\Sigma}$. It exposes only the magnitude of the
  $\ell_1$ penalty `alpha`, which we tune per dataset to achieve a total sparsity
  $\sum_t k_t$ close to the target. This method tends to return PCs with very unequal sparsity
  levels, the first typically denser than the rest.
- **`mixOmics::spca()`** --- takes the data matrix and the list of per-component budgets
  `keepX`, so it hits the cardinality target exactly.
- **`nsprcomp::nsprcomp()`** and **`nsprcomp::nscumcomp()`** --- two functions from the same
  package, both based on the EM algorithm of @sigg2008expectation, which behave differently
  enough that we report them separately. `nsprcomp()` computes the leading sparse PC then
  applies the deflation procedure of @mackey2008deflation for subsequent PCs; it takes the data
  matrix and a vector of per-component cardinalities. `nscumcomp()` solves for all $r$ PCs
  simultaneously under a single cumulative budget $\sum_t k_t$, which it distributes as it sees
  fit, plus the penalty `gamma` on divergence from orthonormality.
- **`amanpg::spca.amanpg()`** --- accepts either input and is told which via `type`: `type = 0`
  declares `z` to be a data matrix, any non-zero value declares it a covariance matrix. We use
  `type = 1` with the correlation matrix on mtcars and Pitprops, where the correlation matrix is
  the natural object, and `type = 0` with the data matrix on breast cancer. It computes all $r$
  components jointly, alternating a proximal-gradient step on the sparse loadings with a manifold
  update of an orthonormal factor on the Stiefel manifold; `lambda1` is a vector of per-component
  $\ell_1$ penalties, which we tune per dataset, and we set `lambda2 = Inf`.
- **`sklearn.decomposition.SparsePCA()`** --- called via `reticulate`. The class computes
  `n_components` components simultaneously using dictionary learning with an $\ell_1$ penalty
  `alpha` on the loadings, applied to the data matrix. We tune `alpha` per dataset and fix
  `random_state` to the repetition's seed.

For Pitprops only the correlation matrix is available, so we generate pseudo-data via
`MASS::mvrnorm()` with the Pitprops matrix as the population covariance ($n = 500$) for the
methods that require $\boldsymbol{X}$ --- `mixOmics::spca()`, `nsprcomp::nsprcomp()`,
`nsprcomp::nscumcomp()`, `PMA::SPC()` and `sklearn...SparsePCA()`. FVE and the orthogonality
violation are always evaluated against the published Pitprops correlation matrix, never against
the covariance of the pseudo-data.

Every tuning parameter is selected by the scripts rather than fixed by hand: each is swept over a
grid and the value whose realised sparsity comes closest to the target is kept, with ties broken
on FVE and candidates that leave a component empty discarded. The selected values appear beneath
each table. Tuning is excluded from all reported runtimes and memory figures.

### How runtime and memory are measured

Every repetition of every method runs in its **own fresh R process**, driven by
[`replication/benchmarking/bench_utils.R`](https://github.com/jeanpauphilet/msPCA/blob/main/replication/benchmarking/bench_utils.R).
Isolation is necessary because peak memory is read as a process-level high-water mark: several
methods in one session, or even several repetitions of one method, contaminate each other.

Memory is peak resident set size, obtained from `getrusage(RUSAGE_SELF).ru_maxrss`. R-level
metrics such as `gc()` or `bench::mark(mem_alloc)` are deliberately **not** used: `msPCA` does
its work in `RcppEigen` buffers and `scikit-learn` in the embedded CPython heap, neither of
which R's garbage collector observes, so an R-level metric would under-report exactly those two
methods while fully charging the pure-R competitors. The **Memory** column below reports the
working set, defined as the size of the matrix handed to the method plus the additional peak
RSS the method allocates on top of it. The R interpreter and package-loading floor are excluded,
so the column is comparable across methods but is not the absolute footprint of the session. It
does include whatever a package allocates on its first call --- lazy loading of namespaces,
compiled code, S4 method tables --- which is on the order of 4 MB.

No warm-up call precedes the measurement, so the ~4 MB of first-call loading is charged to every
method. That is deliberate. Any allocation beforehand leaves R's garbage collector with higher
heap growth targets, so the measured run collects less often and peaks higher, inflating the
methods that allocate on R's heap while leaving `msPCA` and `scikit-learn` untouched. Carrying
the loading cost is the smaller distortion.

Each method is repeated five times with a different seed. Runtime, memory, FVE and the
orthogonality violation are reported as medians, with a bracketed range wherever the spread
exceeds 15%.

Those ranges matter, because neither runtime nor peak RSS is reproducible to the precision a bare
median implies. Wall-clock time varies with machine load and with seed-dependent iteration counts.
Peak RSS varies with garbage-collector timing: repeated runs of the Pitprops script, at identical
seeds and with bit-identical FVE and orthogonality, put `nscumcomp`'s working set anywhere between
72 and 211 MB. For the pure-R methods even the median of five moves by a factor of two between
sittings, so read their memory figures as order-of-magnitude. `msPCA` is the exception: it
allocates in `Eigen` rather than on R's heap, and its figures reproduce to within a few percent.

```{r}
# Representative call pattern; see replication/benchmarking/ for the full scripts.
set.seed(42)
res <- mspca(Sigma, r = 3, ks = rep(4, 3), verbose = FALSE)

fraction_variance_explained(Sigma, res$x_best)
feasibility_violation_off(Sigma, res$x_best, 0)
res$runtime
```

## Results

### mtcars

`datasets::mtcars` from base R ($p = 11$ variables, $n = 32$ observations) serves as a simple
reproducible baseline. We extract $r = 3$ sparse PCs targeting $k = 4$ per component.

| Method | $(k_1,k_2,k_3)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (4,4,4) | 0.829 | 9.4e-05 | 0.193 | 0.4 [0.4, 0.6] |
| `msPCA::mspca` $(\boldsymbol{X})$ | (4,4,4) | 0.829 | 9.4e-05 | 0.245 | 0.2 [0.2, 0.5] |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (4,4,4) | 0.753 | 0.117 | 0.092 | 21 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (4,4,4) | 0.754 | 0.117 | 0.112 | 23 |
| `PMA::SPC` | (4,4,5) | 0.551 | 0.221 | 0.009 | 0.6 |
| `sparsepca::spca` | (6,4,1) | 0.581 | 0.133 | 0.031 | 17 |
| `mixOmics::spca` | (4,4,4) | 0.633 | 0.133 | 0.003 [0.003, 0.004] | 0.5 |
| `nsprcomp::nsprcomp` | (4,4,4) | 0.832 | 0.078 | 0.004 [0.004, 0.005] | 0.1 [0.1, 1.2] |
| `nsprcomp::nscumcomp` | (5,5,2) | 0.824 | 1.1e-05 | 0.242 | 30 [30, 35] |
| `amanpg::spca.amanpg` | (4,4,1) | 0.544 | 0.047 | 0.008 | 1.6 |
| `sklearn...SparsePCA` | (4,3,4) | 0.827 | < 1e-15 | 0.013 [0.009, 0.015] | 0.0 [0.0, 0.1] |
| PCA (dense) | (11,11,11) | 0.899 | < 1e-15 | 0.001 [0.001, 0.002] | 0.5 |

Tuning selected automatically: `PMA::SPC` `sumabsv = 1.85`; `sparsepca::spca` `alpha = 0.0045`;
`amanpg::spca.amanpg` `lambda1 = c(4.5, 5.4, 0.1)`; `sklearn...SparsePCA` `alpha = 3.2`;
`nsprcomp::nscumcomp` `gamma = 1e6`. Thirty-one of the forty `sparsepca` candidates and five of the
`sklearn` candidates were rejected for leaving a component empty.

`nscumcomp` performs well here once `gamma` is tuned. At 0.824 it is the second-best sparse
method by FVE, and at 1.1e-05 one of the few inside the feasibility tolerance, though it returns
(5,5,2) rather than the requested (4,4,4). At its default `gamma = 0` the same call gives a
violation of 0.242.

Only `nscumcomp` varies across seeds, over 0.798--0.825 in FVE, with a worst-case violation of
3.3e-05 that remains inside tolerance. Every other method, `msPCA` included, returns the same
solution at every seed.

At $p = 11$ the numerical working set of every method is a few kilobytes, so the memory column
measures fixed overheads rather than algorithmic cost. Most methods sit at or below 1.6 MB.
`elasticnet`, `sparsepca` and `nscumcomp` do not: each carries a floor of 17--30 MB already
present at this size, of which only about 4 MB is first-call loading.

### Pitprops

The Pitprops dataset [@jeffers1967two] ($p = 13$ physical measurements on $n = 180$ timber
specimens) is a classic sparse PCA benchmark [e.g. @zou2006sparse]. We extract $r = 6$ sparse
PCs targeting $k = 4$ nonzero loadings per component, as in @zou2006sparse.

| Method | $(k_1, \ldots, k_6)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (4,4,4,4,4,4) | 0.791 | 9.7e-05 | 0.650 | 0.6 [0.4, 0.8] |
| `msPCA::mspca` $(\boldsymbol{X})$ | (4,4,4,4,4,4) | 0.794 | 9.4e-05 | 3.972 | 0.5 [0.3, 0.5] |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (4,4,4,4,4,4) | 0.792 | 0.241 | 0.070 | 21 [21, 27] |
| `elasticnet::spca` $(\boldsymbol{X})$ | (4,4,4,4,4,4) | 0.790 | 0.322 | 0.158 | 78 [70, 83] |
| `PMA::SPC` | (3,4,4,4,3,4) | 0.714 | 0.491 | 0.020 [0.020, 0.024] | 15 |
| `sparsepca::spca` | (7,3,4,2,4,4) | 0.756 | 0.353 | 0.039 | 28 |
| `mixOmics::spca` | (4,4,4,4,4,4) | 0.788 | 0.215 | 0.009 | 0.6 |
| `nsprcomp::nsprcomp` | (4,4,4,4,4,4) | 0.816 | 0.563 | 0.025 [0.023, 0.028] | 33 [32, 39] |
| `nsprcomp::nscumcomp` | (5,6,5,2,4,2) | 0.819 | 5.1e-05 | 1.086 | 65 [64, 129] |
| `amanpg::spca.amanpg` | (3,3,4,3,4,6) | 0.863 | 1.654 | 0.011 | 4.5 [4.5, 7.2] |
| `sklearn...SparsePCA` | (7,3,4,3,5,4) | 0.907 | 0.815 | 0.061 | 0.1 [0.1, 0.4] |
| PCA (dense) | (13,...,13) | 0.877 | < 1e-15 | 0.001 | 0.4 |

Tuning selected automatically: `PMA::SPC` `sumabsv = 1.55`; `sparsepca::spca` `alpha = 0.005`;
`amanpg::spca.amanpg` `lambda1 = c(4.80, 4.40, 1.95, 5.50, 0.20, 0.05)`; `sklearn...SparsePCA`
`alpha = 2.6`; `nsprcomp::nscumcomp` `gamma = 1e7`. The spread in `amanpg`'s penalty vector, from
5.50 down to 0.05, tracks the decay of the spectrum; a single shared value cannot reproduce it.

`nscumcomp` needs $\gamma = 10^7$, seven orders of magnitude above its default, to bring its
orthogonality violation to 5.1e-05 from the 1.87 it returns at $\gamma = 0$. That costs 0.026 of
FVE and a fortyfold slowdown, and the resulting sparsity is (5,6,5,2,4,2) rather than the
requested (4,4,4,4,4,4).

Across seeds, `msPCA` $(\boldsymbol{\Sigma})$ ranges over 0.786--0.822 in FVE and `msPCA`
$(\boldsymbol{X})$ over 0.784--0.801, both holding the orthogonality violation below 1e-04 at every
seed. `nsprcomp` ranges over 0.813--0.833 with a worst-case violation of 1.105, and `nscumcomp`
over 0.814--0.830 with a worst case of 1.03e-04. This is the one dataset where `msPCA` is visibly
stochastic, an artefact of the random restarts in the truncated power method; feasibility is what
does not vary.

This is also the dataset where the feasibility differences are starkest. With $p = 13$ and a total
budget of 24 nonzeros, trivially orthogonal disjoint-support solutions do not exist, so methods
that rely on deflation have nothing to fall back on: at the requested cardinality, `elasticnet`,
`mixOmics` and `nsprcomp` return violations of 0.215--0.563 against 9.7e-05 for `msPCA`. Only
`nscumcomp`, with `gamma` driven to $10^7$, also reaches the tolerance --- and it does so at a
cardinality that misses the budget.

Pitprops is the one case where `type = "X"` is the wrong choice: with $n = 500$ pseudo-observations
and $p = 13$, the data matrix is larger than the correlation matrix, and the $O(np)$ matrix-free
product is more expensive than the $O(p^2)$ dense one. Runtime is roughly six times worse than
the `type = "Sigma"` path, though at this scale the memory difference is below what the
measurement can resolve. The interface is a win when $n \ll p$, not universally.

### Breast cancer gene expression

Gene expression data from @chin2006genomic ($n = 89$ tumor samples, 19,672 genes). Following
the preprocessing of @witten2009penalized we retain the $p = 500$ genes with the highest
marginal variance, yielding a high-dimensional ($p > n$) benchmark. We set $r = 3$ and target
$k = 20$ nonzero loadings per component.

| Method | $(k_1, k_2, k_3)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (20,20,20) | 0.093 | 7.5e-05 | 27.208 [14.083, 27.549] | 7.1 [6.5, 9.0] |
| `msPCA::mspca` $(\boldsymbol{X})$ | (20,20,20) | 0.095 | 7.5e-05 | 13.232 [6.808, 13.320] | 3.1 [2.9, 4.2] |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (20,20,20) | 0.043 | 0.040 | 16.180 | 209 [202, 255] |
| `elasticnet::spca` $(\boldsymbol{X})$ | (20,20,20) | 0.043 | 0.021 | 0.812 | 79 [77, 152] |
| `PMA::SPC` | (20,22,13) | 0.042 | 0.008 | 0.042 | 34 [33, 39] |
| `sparsepca::spca` | (43,14,22) | 0.082 | < 1e-15 | 4.770 | 92 [49, 124] |
| `mixOmics::spca` | (20,20,20) | 0.077 | < 1e-15 | 0.032 | 7.0 |
| `nsprcomp::nsprcomp` | (20,20,20) | 0.083 | < 1e-15 | 0.033 [0.030, 0.037] | 48 [48, 58] |
| `nsprcomp::nscumcomp` | (29,23,8) | 0.081 | < 1e-15 | 0.232 [0.220, 0.264] | 57 |
| `amanpg::spca.amanpg` | (68,1,2) | 0.045 | < 1e-15 | 0.071 | 50 |
| `sklearn...SparsePCA` | (22,3,1) | 0.020 | < 1e-15 | 0.079 | 0.6 [0.6, 0.7] |
| PCA (dense) | (500,500,500) | 0.352 | < 1e-15 | 0.010 | 2.6 |

Tuning selected automatically: `PMA::SPC` `sumabsv = 3.3`; `sparsepca::spca` `alpha = 0.0035`;
`amanpg::spca.amanpg` `lambda1 = c(4.5, 1.8, 3.3)`; `sklearn...SparsePCA` `alpha = 15`;
`nsprcomp::nscumcomp` `gamma = 316`, with a total budget of $k = 60$ that it distributes unevenly.
`gamma` matters far less here than on the small datasets, because at $p = 500$ with 60 nonzeros in
total the supports are nearly disjoint anyway.

This is the dataset where the penalty-parameterized methods break down most clearly. No value on
the grid lets `amanpg::spca.amanpg()` approach (20,20,20): the best available is (68,1,2), a
deviation of 85, and the two trailing components are all but empty. `sklearn...SparsePCA()`
reaches (22,3,1), and `sparsepca::spca()` (43,14,22). Only the four functions that accept an
exact cardinality --- `msPCA`, `elasticnet`, `mixOmics` and `nsprcomp::nsprcomp` --- return the
budget that was asked for.

Across seeds, `msPCA` ranges over 0.088--0.095 in FVE in both input modes, with the orthogonality
violation at or below 9.8e-05 at every seed. `nsprcomp` ranges over 0.070--0.084, `nscumcomp` over
0.080--0.085, and `sklearn...SparsePCA` over 0.020--0.032.

With $p = 500$ and 60 nonzeros in total, disjoint supports are easy to find, so most methods are
trivially orthogonal here and the feasibility column carries little information.

This is the smallest dataset at which the memory column measures the algorithms rather than fixed
overheads. `msPCA` is the lightest of the sparse methods that meet the budget: 7.1 MB on the
correlation matrix and 3.1 MB on the data matrix, against 209 MB for `elasticnet` in Gram mode,
79 MB on the data matrix, 92 MB for `sparsepca`, and 48--57 MB for `nsprcomp` and `nscumcomp`.
`mixOmics` at 7.0 MB is comparable, and `sklearn` lighter still at 0.6 MB, though it allocates
outside R entirely and returns (22,3,1) rather than the requested budget. The `type = "X"` saving
is modest here because $\boldsymbol{\Sigma}$ is only about 2 MB at $p = 500$. Riboflavin is where it
becomes decisive.

### Riboflavin

The riboflavin (vitamin B2) production dataset [@buhlmann2014high] comprises log-transformed
expression levels of $p = 4{,}088$ genes in $n = 71$ samples of *Bacillus subtilis*. This is a
challenging $p \gg n$ benchmark: the empirical correlation matrix is rank-deficient
(rank $\leq 70$). We set $r = 2$ and target $k = 20$ nonzero loadings per component. To avoid
time-consuming parameter tuning, we consider only methods that accept exact cardinality
budgets.

| Method | $(k_1, k_2)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (20,20) | 0.008 | < 1e-15 | 59.504 [50.630, 99.328] | 641 |
| `msPCA::mspca` $(\boldsymbol{X})$ | (20,20) | 0.008 | < 1e-15 | 1.842 [0.891, 2.563] | 45 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (20,20) | 0.004 | < 1e-15 | 6.894 | 309 [307, 426] |
| `mixOmics::spca` | (20,20) | 0.006 | < 1e-15 | 0.334 | 24 [20, 24] |
| `nsprcomp::nsprcomp` | (20,20) | 0.009 | < 1e-15 | 0.353 [0.266, 0.371] | 59 |
| PCA (dense) | (4088,4088) | 0.478 | < 1e-15 | 0.046 | 39 |

`msPCA` is run with `maxIter = 100` here; all methods other than `msPCA` $(\boldsymbol{\Sigma})$ use the
data matrix directly. Only `msPCA` varies at all across seeds, and only in the third decimal of
FVE (0.008--0.009 in both input modes).

This dataset isolates the benefit of the `type = "X"` interface. Passing the raw data matrix
replaces the $O(p^2)$ product $\boldsymbol{\Sigma}\boldsymbol{\beta}$ with the $O(np)$ two-pass product
$\boldsymbol{X}^\top(\boldsymbol{X}\boldsymbol{\beta})/(n-1)$ and never forms the covariance matrix. The measured working
set falls from 641 MB to 45 MB, a factor of about 14, at identical FVE and feasibility. The
$p \times p$ correlation matrix accounts for 128 MB of that, the remainder being the `Eigen` copy
of it and the dense intermediates the $O(p^2)$ path requires. Median runtime falls from 59.5
seconds to 1.8, a factor of about 30, though both rows vary by a factor of two or more across
seeds and one significant figure is all the ratio will support.

The memory figures over those same repetitions span 639--642 MB, a spread of well under one
percent. The contrast holds throughout: `msPCA` allocates in `Eigen`, where the footprint follows
from the algorithm, so its memory requirement can be predicted from $n$, $p$ and $r$. The pure-R
methods allocate on R's heap, where the peak depends on garbage-collector timing, and
`elasticnet`'s median on this dataset has landed anywhere between 309 and 570 MB on repeated runs
of the identical script.

Together with the Pitprops result the rule is simply the relative size of the two inputs: use
`type = "X"` when $n \ll p$, and `type = "Sigma"` when $p \lesssim n$.

## Discussion

**Sparsity control.** `PMA::SPC()`, `sparsepca::spca()`, `nsprcomp::nscumcomp()`,
`amanpg::spca.amanpg()` and `sklearn.decomposition.SparsePCA()` give no direct control over the
sparsity of each loading vector. `PMA::SPC()` and `sklearn` expose only a global $\ell_1$ bound;
`sparsepca::spca()`, `nscumcomp()` and `amanpg` tend to return unbalanced components, with one PC
carrying most of the budget. None can be calibrated to an exact per-component target, which is
what makes an apples-to-apples comparison awkward. Where the discussion below needs methods that
meet the budget exactly, those are `msPCA::mspca()`, `elasticnet::spca()`, `mixOmics::spca()` and
`nsprcomp::nsprcomp()`.

**Feasibility.** Two things have to be separated here, because the answer depends on whether
disjoint supports exist at all. On breast cancer and riboflavin, where $p$ is large relative to
the total budget, orthogonality is nearly free: most methods return exactly orthogonal loadings
simply because their supports do not overlap. On mtcars ($p = 11$, 12 nonzeros) and Pitprops
($p = 13$, 24 nonzeros) no disjoint solution exists, and that is where the differences appear.

There, `msPCA::mspca()` reaches 1e-4 while returning the requested cardinality exactly, and does
so at every repetition: the largest violation it produced anywhere in the benchmark is 9.9e-05.
Two other methods reach the tolerance on one dataset each, both with qualifications.
`nsprcomp::nscumcomp()` reaches it on both, but only with `gamma` driven to 10^6--10^7, returning
(5,5,2) and (5,6,5,2,4,2) against budgets of (4,4,4) and (4,4,4,4,4,4), and its worst Pitprops
repetition is 1.03e-04. `sklearn...SparsePCA()` reaches it on mtcars by returning 11 nonzeros in
total rather than the requested 12, which at $p = 11$ admits a disjoint solution and so makes
orthogonality free. Every other method, at the requested cardinality, returns violations between
0.08 and 0.56 on these two datasets.

The claim supported by this is narrower than "only `msPCA` can be feasible", and more useful: it
is the only method that delivers feasibility and the requested sparsity together, without a search
over a penalty parameter to get there.

**Variance explained.** Among the methods that meet the requested cardinality,
`msPCA::mspca()` and `nsprcomp::nsprcomp()` are generally the best performers, followed by
`mixOmics::spca()` and `elasticnet::spca()`. `msPCA` does not appear to sacrifice variance
explained in order to achieve feasibility.

Several methods post higher FVE while missing the budget, and those figures should be read in
that light: `sklearn...SparsePCA()` reaches 0.907 on Pitprops at (7,3,4,3,5,4) with a violation of
0.815, and `amanpg::spca.amanpg()` 0.863 at a violation of 1.654. The exception is `nscumcomp` at
0.824 on mtcars with (5,5,2), which is both high and feasible, and is the most substantive
competition in the comparison.

**Reproducibility.** Repeating each method under several seeds separates the deterministic
implementations (`elasticnet`, `PMA`, `sparsepca`, `mixOmics`, `amanpg`, dense PCA) from the rest.
`nsprcomp` and `nscumcomp` use random initialisation and move between runs; `nsprcomp`'s
orthogonality violation on Pitprops spans 0.56--1.10. `msPCA` uses random restarts and is likewise
not bit-reproducible across seeds, but its variation is confined to the objective. The feasibility
guarantee holds at every seed on every dataset, the largest violation observed anywhere being
9.9e-05.

**Memory.** On the two high-dimensional datasets, where the memory column reflects the
algorithms rather than fixed overheads, `msPCA` is the lightest of the sparse methods that
respect an exact cardinality budget. On breast cancer it uses 3.1 MB on the data matrix against
79 MB for `elasticnet` on the same input and 209 MB in Gram mode, with `sparsepca` at 92 MB and
`nsprcomp` and `nscumcomp` at 48--57 MB. On riboflavin it uses 45 MB against 309 MB for
`elasticnet`. Within `msPCA`, the $\boldsymbol{\Sigma}$-versus-$\boldsymbol{X}$ choice matters more than the
choice of package does between packages: the same function spans 641 MB to 45 MB on riboflavin
depending on the interface.

Those figures are also more stable than the competitors', for the reason given in the riboflavin
section. For anyone sizing a job, a predictable memory requirement is worth as much as a small
one, and the competing figures quoted here should be read as order-of-magnitude.

**Scalability.** `msPCA::mspca()` scales to datasets with $p$ in the thousands in reasonable
time, and the `type = "X"` interface yields substantial runtime and memory gains when
$n \ll p$: roughly 30x faster and 14x lighter on riboflavin. When $n > p$, as in the Pitprops
pseudo-data, the ordering reverses and `type = "Sigma"` is preferable.

## References
