---
title: "Benchmarking against other sparse PCA packages"
output: rmarkdown::html_vignette
bibliography: thebib.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette gives the full benchmarking protocol and results summarised in the package
description. It compares `msPCA` against seven competing implementations on four real
datasets, and documents exactly how each competitor was configured --- which matters, because
the packages parameterise sparsity in incompatible ways.

Chunks are not evaluated at build time: reproducing the full comparison requires eight
additional packages, `callr` for the benchmarking harness, and a Python installation. The
memory measurements additionally require macOS or Linux. The complete scripts live in the
package repository under
[`replication/benchmarking/`](https://github.com/jeanpauphilet/msPCA/tree/main/replication/benchmarking).

## Getting the data

All four datasets are obtainable from R without manual downloads or accounts, so the
comparison is reproducible end to end:

```{r}
## mtcars -- base R
X_mtcars     <- as.matrix(datasets::mtcars)
Sigma_mtcars <- cor(X_mtcars)

## Pitprops -- the 13 x 13 correlation matrix of Jeffers (1967), Table 1.
## It is transcribed literally at the top of
## replication/benchmarking/notebook_pitprops.R; source that file to obtain `pitprops`.
Sigma_pitprops <- pitprops
## Methods that require a data matrix instead use pseudo-data drawn from this
## correlation structure (see "How each method was configured" below).
set.seed(42)
X_pitprops <- MASS::mvrnorm(n = 500, mu = rep(0, ncol(pitprops)),
                            Sigma = pitprops)

## Breast cancer -- one-off download via PMA, then keep the 500 highest-
## variance genes, following the preprocessing of Witten et al. (2009).
X            <- t(PMA::download_breast_data(
  url = "https://tibshirani.su.domains/PMA/breastdata.rda")$rna)
X_breast     <- X[, order(apply(X, 2, var), decreasing = TRUE)[1:500]]
Sigma_breast <- cor(X_breast)          # n = 89, p = 500

## Riboflavin -- ships with the hdi package
data("riboflavin", package = "hdi")
X_ribo <- as.matrix(riboflavin$x)      # n = 71, p = 4088
```

Only the breast cancer data touches the network, and `PMA::download_breast_data()` handles it
in a single call. Riboflavin ships inside `hdi`, `mtcars` is in base R, and the Pitprops
correlation matrix is transcribed in the benchmarking script, so no manual downloads or
accounts are needed anywhere.

## Methods compared

We compare `msPCA` against seven competing methods:

| Package | Reference | Language | Approach |
|---|---|---|---|
| `elasticnet` | @zou2006sparse | R | elastic-net regularization, LARS-like |
| `PMA` | @witten2009penalized | R | penalized matrix decomposition |
| `sparsepca` | @erichson2020sparse | R | variable projection with $\ell_1$ penalty |
| `mixOmics` | @rohart2017mixomics | R | regularized SVD [@shen2008sparse] |
| `nsprcomp` | @sigg2019nsprcomp | R | expectation-maximization [@sigg2008expectation] + deflation |
| `amanpg` | @chen2020proximal | R | alternating manifold proximal gradient |
| `sklearn.decomposition.SparsePCA` | @pedregosa2011scikit | Python | dictionary learning |

`scikit-learn` is called from R via the `reticulate` package [@ushey2023reticulate]. Together
these cover the main methodological approaches to sparse PCA: elastic-net regularization,
penalized matrix decomposition, variable-projection optimization, regularized SVD, deflation,
Riemannian proximal gradient, and dictionary learning.

## Environment

Experiments were conducted on a MacBook Air with an Apple M2 chip and 24 GB of RAM, with
R version 4.6.0 (2026-04-24), `msPCA` v0.5.0, `elasticnet` v1.3, `PMA` v1.2.4, `sparsepca`
v0.1.2, `mixOmics` v6.36.0, `nsprcomp` v0.5.1.2, `amanpg` v0.1.0, `reticulate` v1.40.0, and
`scikit-learn` v1.6.1 (Python 3.11). Full session information is recorded in
[`replication/sessionInfo.txt`](https://github.com/jeanpauphilet/msPCA/blob/main/replication/sessionInfo.txt).

## How each method was configured

Comparing sparse PCA methods is complicated because different packages parameterize sparsity
differently. Some accept a covariance or correlation matrix, others apply only to the data
matrix, and some support both.

- **`msPCA`** --- `mspca()` with both `type = "Sigma"` and `type = "X"` on all four datasets.
  On riboflavin ($n \ll p$) the $p \times p$ covariance matrix is rank-deficient, making
  `type = "X"` the natural choice; we report both modes to quantify the runtime gain.
- **`elasticnet`** --- `spca()`, which takes a list of cardinality budgets (one per PC) and
  applies interchangeably to $\boldsymbol{\Sigma}$ or $\boldsymbol{X}$. We report both input modes on mtcars,
  Pitprops and breast cancer.
- **`PMA`** --- `SPC()` with `orth = TRUE`. It takes the covariance matrix and a bound
  `sumabsv` on $\sum_t \|\boldsymbol{u}_t\|_1$, which we tune manually per dataset to get
  per-component sparsity as close as possible to the target.
- **`sparsepca`** --- `spca()` applied to $\boldsymbol{\Sigma}$, which only exposes the magnitude of
  the $\ell_1$ penalty `alpha`. We tune it per dataset to achieve a total sparsity
  $\sum_t k_t$ close to the target. This method tends to return PCs with very unequal sparsity
  levels, the first typically denser than the rest.
- **`mixOmics`** --- `spca()`, which takes the data matrix and the list of sparsity budgets.
- **`nsprcomp`** --- two functions, both based on the EM algorithm of @sigg2008expectation.
  `nsprcomp()` computes the leading sparse PC then applies the deflation procedure of
  @mackey2008deflation for subsequent PCs; it takes the data matrix and a list of sparsity
  levels. `nscumcomp()` solves for all $r$ PCs simultaneously, taking a total sparsity budget
  $\sum_t k_t$ and a penalty parameter `gamma` on orthogonality violation (default 0).
- **`amanpg`** --- `spca.amanpg()` applied to the correlation matrix (`type = 1`). It solves a
  single-component problem with an $\ell_1$ penalty `lambda1` and applies sequential deflation;
  we tune `lambda1` per dataset.
- **`scikit-learn`** --- `sklearn.decomposition.SparsePCA()` via `reticulate`. The class
  computes $r$ components simultaneously using dictionary learning with an $\ell_1$ penalty
  `alpha` on the loadings, applied to the data matrix. We tune `alpha` per dataset.

For Pitprops only the correlation matrix is available, so we generate pseudo-data via
`MASS::mvrnorm()` with the Pitprops matrix as the population covariance ($n = 500$) for the
methods that require $\boldsymbol{X}$.

For each method we report the actual number of nonzero loadings per PC, the FVE computed on
the original correlation matrix, the orthogonality violation, the runtime, and the memory
footprint. Hyperparameter tuning is excluded from all reported figures.

### How runtime and memory are measured

Every repetition of every method runs in its **own fresh R process**, driven by
[`replication/benchmarking/bench_utils.R`](https://github.com/jeanpauphilet/msPCA/blob/main/replication/benchmarking/bench_utils.R).
Isolation is necessary because peak memory is read as a process-level high-water mark: several
methods in one session, or even several repetitions of one method, contaminate each other.

Memory is peak resident set size, obtained from `getrusage(RUSAGE_SELF).ru_maxrss`. R-level
metrics such as `gc()` or `bench::mark(mem_alloc)` are deliberately **not** used: `msPCA` does
its work in `RcppEigen` buffers and `scikit-learn` in the embedded CPython heap, neither of
which R's garbage collector observes, so an R-level metric would under-report exactly those two
methods while fully charging the pure-R competitors. The **Memory** column below reports the
working set, defined as the size of the matrix handed to the method plus the additional peak
RSS the method allocates on top of it. The R interpreter and package-loading floor are excluded,
so the column is comparable across methods but is not the absolute footprint of the session. It
does include whatever a package allocates on its first call --- lazy loading of namespaces,
compiled code, S4 method tables --- which is on the order of 4 MB.

Each method is repeated five times with a **different seed per repetition**. Runtime, memory,
FVE and the orthogonality violation are all reported as medians over the five.

The memory column additionally carries a range, because **peak RSS is not reproducible**, even
for deterministic code: it depends on when the garbage collector happens to run. Two identical
runs of the Pitprops benchmark, at identical seeds and with bit-identical FVE and orthogonality,
put `nscumcomp`'s working set at 72 MB and then at 211 MB. Where the bracketed range beneath a
memory figure is wide, the median should not be leaned on. The ranges are narrow for `msPCA`,
which does its allocation in Eigen rather than on R's heap, and widest for the methods that
allocate most heavily in R.

An earlier version of this benchmark called each method once on a small problem before the
measurement, to move first-call loading cost out of the memory column. That was removed. Any
allocation before the measurement leaves R's garbage collector with higher heap growth targets,
so the measured run collects less often and peaks higher --- inflating precisely those methods
that allocate on R's heap while leaving `msPCA` and `scikit-learn` untouched, and so biasing the
comparison in `msPCA`'s favour. Carrying the ~4 MB of loading cost in every row is the lesser
distortion.

```{r}
# Representative call pattern; see replication/benchmarking/ for the full scripts.
set.seed(42)
res <- mspca(Sigma, r = 3, ks = rep(4, 3), verbose = FALSE)

fraction_variance_explained(Sigma, res$x_best)
feasibility_violation_off(Sigma, res$x_best, 0)
res$runtime
```

## Results

### mtcars

`datasets::mtcars` from base R ($p = 11$ variables, $n = 32$ observations) serves as a simple
reproducible baseline. We extract $r = 3$ sparse PCs targeting $k = 4$ per component.

| Method | $(k_1,k_2,k_3)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (4,4,4) | 0.829 | 9.4e-05 | 0.189 | 0.4 [0.4, 0.6] |
| `msPCA::mspca` $(\boldsymbol{X})$ | (4,4,4) | 0.829 | 9.4e-05 | 0.244 | 0.2 [0.2, 0.4] |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (4,4,4) | 0.753 | 0.117 | 0.092 | 21 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (4,4,4) | 0.754 | 0.117 | 0.111 | 23 |
| `PMA::SPC` | (4,4,5) | 0.552 | 0.283 | 0.006 | 0.6 |
| `sparsepca::spca` | (7,4,1) | 0.585 | 0.045 | 0.032 | 17 |
| `mixOmics::spca` | (4,4,4) | 0.633 | 0.133 | 0.003 | 0.5 |
| `nsprcomp::nsprcomp` | (4,4,4) | 0.832 | 0.078 | 0.004 | 0.1 |
| `nsprcomp::nscumcomp` | (6,3,3) | 0.772 | 0.242 | 0.024 | 16 [14, 20] |
| `amanpg::spca.amanpg` | (7,4,2) | 0.649 | 0.041 | 0.003 | 0.5 |
| `sklearn...SparsePCA` | (4,3,4) | 0.827 | < 1e-15 | 0.017 | 0.1 [0.1, 0.2] |
| PCA (dense) | (11,11,11) | 0.899 | < 1e-15 | 0.001 | 0.5 |

Tuning: `PMA` uses `sumabsv = 1.69`; `sparsepca` uses `alpha = 0.004`; `amanpg` uses
`lambda1 = c(10, 3, 0.01)`; `scikit-learn` uses `alpha = 3.2`.

Stability across seeds: only `nscumcomp` varies, with FVE ranging over 0.753--0.823 and the
orthogonality violation reaching 0.289 in the worst run. Every other method, `msPCA` included,
returns the same solution at every seed.

At $p = 11$ the numerical working set of every method is a few kilobytes, so the memory column
here measures fixed overheads rather than algorithmic cost. Most methods sit at or below 0.6 MB,
as expected. `elasticnet`, `sparsepca` and `nscumcomp` do not: they carry a floor of 16--23 MB
that is already present at this size. A single-repetition check confirmed the floor is genuine
allocation, not first-call loading, which accounts for only about 4 MB of it.

### Pitprops

The Pitprops dataset [@jeffers1967two] ($p = 13$ physical measurements on $n = 180$ timber
specimens) is a classic sparse PCA benchmark [e.g. @zou2006sparse]. We extract $r = 6$ sparse
PCs targeting $k = 4$ nonzero loadings per component, as in @zou2006sparse.

| Method | $(k_1, \ldots, k_6)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (4,4,4,4,4,4) | 0.791 | 9.7e-05 | 0.646 | 0.4 [0.4, 0.8] |
| `msPCA::mspca` $(\boldsymbol{X})$ | (4,4,4,4,4,4) | 0.794 | 9.4e-05 | 3.953 | 0.3 [0.3, 0.8] |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (4,4,4,4,4,4) | 0.792 | 0.241 | 0.070 | 21 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (4,4,4,4,4,4) | 0.790 | 0.322 | 0.158 | 83 [70, 83] |
| `PMA::SPC` | (4,4,4,7,6,4) | 0.660 | 0.857 | 0.014 | 6.8 [6.6, 7.5] |
| `sparsepca::spca` | (8,4,3,1,3,5) | 0.762 | 0.376 | 0.026 | 16 [16, 17] |
| `mixOmics::spca` | (4,4,4,4,4,4) | 0.788 | 0.215 | 0.009 | 0.6 |
| `nsprcomp::nsprcomp` | (4,4,4,4,4,4) | 0.816 | 0.563 | 0.025 | 32 [28, 39] |
| `nsprcomp::nscumcomp` | (5,4,2,6,4,3) | 0.845 | 1.874 | 0.346 | 66 [63, 69] |
| `amanpg::spca.amanpg` | (8,6,9,2,3,5) | 0.760 | 1.763 | 0.010 | 3.7 |
| `sklearn...SparsePCA` | (7,3,4,2,4,4) | 0.895 | 0.733 | 0.045 | 0.1 |
| PCA (dense) | (13,...,13) | 0.877 | < 1e-15 | < 0.001 | 0.4 |

Tuning: `PMA` uses `sumabsv = 1.727`; `sparsepca` uses `alpha = 0.004`; `nscumcomp` uses the
default `gamma = 0`; `amanpg` uses `lambda1 = c(1, 1, 5, 0.4, 0.4, 0)`; `scikit-learn` uses
`alpha = 3.1`.

Stability across seeds: `msPCA` $(\boldsymbol{\Sigma})$ ranges over 0.786--0.822 in FVE and `msPCA`
$(\boldsymbol{X})$ over 0.784--0.801, in both cases holding the orthogonality violation below 1e-04 at
every seed. `nsprcomp` ranges over 0.813--0.833 with a worst-case violation of 1.105, and
`nscumcomp` over 0.750--0.905 with a worst case of 2.704. This is the one dataset where `msPCA`
is visibly stochastic, an artefact of the random restarts in the truncated power method; what
does not vary is feasibility.

This is also the dataset where the feasibility differences are starkest: competitors return
orthogonality violations in the 0.2--1.9 range, against 9.7e-05 for `msPCA`. With $p = 13$ and
a total budget of 24 nonzeros, trivially orthogonal disjoint-support solutions do not exist,
so methods that rely on deflation have nothing to fall back on. Note also that `nscumcomp`
reports the highest FVE of any sparse method here (0.845), but does so while violating
orthogonality by 1.9 --- the two numbers have to be read together.

Pitprops is the one case where `type = "X"` is the wrong choice: with $n = 500$ pseudo-observations
and $p = 13$, the data matrix is larger than the correlation matrix, and the $O(np)$ matrix-free
product is more expensive than the $O(p^2)$ dense one. Runtime is six times worse than the
`type = "Sigma"` path, though at this scale the memory difference is below what the measurement
can resolve. The interface is a win when $n \ll p$, not universally.

### Breast cancer gene expression

Gene expression data from @chin2006genomic ($n = 89$ tumor samples, 19,672 genes). Following
the preprocessing of @witten2009penalized we retain the $p = 500$ genes with the highest
marginal variance, yielding a high-dimensional ($p > n$) benchmark. We set $r = 3$ and target
$k = 20$ nonzero loadings per component.

| Method | $(k_1, k_2, k_3)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (20,20,20) | 0.093 | 7.5e-05 | 27.257 | 6.6 [6.3, 9.5] |
| `msPCA::mspca` $(\boldsymbol{X})$ | (20,20,20) | 0.095 | 7.5e-05 | 13.280 | 3.0 [2.8, 3.9] |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (20,20,20) | 0.043 | 0.040 | 18.045 | 282 [265, 320] |
| `elasticnet::spca` $(\boldsymbol{X})$ | (20,20,20) | 0.043 | 0.021 | 0.993 | 174 [133, 232] |
| `PMA::SPC` | (14,17,27) | 0.028 | 0.001 | 0.053 | 38 [35, 45] |
| `sparsepca::spca` | (46,13,8) | 0.061 | < 1e-15 | 4.978 | 56 [45, 169] |
| `mixOmics::spca` | (20,20,20) | 0.077 | < 1e-15 | 0.033 | 7.0 [6.0, 7.1] |
| `nsprcomp::nsprcomp` | (20,20,20) | 0.083 | < 1e-15 | 0.034 | 48 [47, 48] |
| `nsprcomp::nscumcomp` | (33,21,6) | 0.081 | < 1e-15 | 0.167 | 53 [52, 57] |
| `amanpg::spca.amanpg` | (96,4,2) | 0.067 | 0.063 | 0.194 | 55 [54, 57] |
| `sklearn...SparsePCA` | (56,3,6) | 0.057 | < 1e-15 | 0.110 | 1.3 |
| PCA (dense) | (500,500,500) | 0.352 | < 1e-15 | 0.010 | 2.6 [2.6, 3.6] |

Tuning: `PMA` uses `sumabsv = 2.92`; `sparsepca` uses `alpha = 0.004` but returns unequal
sparsity levels; `nscumcomp` receives a total budget of $k = 60$ and distributes it unevenly;
`amanpg` uses `lambda1 = c(8.5, 2.5, 1.5)`; `scikit-learn` uses `alpha = 11.25`.

Stability across seeds: `msPCA` ranges over 0.088--0.095 in FVE in both input modes, with the
orthogonality violation at or below 9.8e-05 at every seed. `nsprcomp` ranges over 0.070--0.084
and `nscumcomp` over 0.078--0.083, the latter reaching a violation of 0.047 in its worst run.

With $p = 500$ and only 60 nonzeros in total, disjoint supports are easy to find, so most
methods are trivially orthogonal here.

This is the smallest dataset at which the memory column measures the algorithms rather than
fixed overheads, and `msPCA` is the lightest sparse method on it by a wide margin: 6.6 MB on the
correlation matrix and 3.0 MB on the data matrix, against 282 MB for `elasticnet` in Gram mode,
174 MB for `elasticnet` on the data matrix, and 48--56 MB for `sparsepca`, `nsprcomp`,
`nscumcomp` and `amanpg`. Only `mixOmics`, at 7.0 MB, is comparable. The `type = "X"` saving is
modest here because $\boldsymbol{\Sigma}$ is only about 2 MB at $p = 500$; riboflavin is where it becomes
decisive.

### Riboflavin

The riboflavin (vitamin B2) production dataset [@buhlmann2014high] comprises log-transformed
expression levels of $p = 4{,}088$ genes in $n = 71$ samples of *Bacillus subtilis*. This is a
challenging $p \gg n$ benchmark: the empirical correlation matrix is rank-deficient
(rank $\leq 70$). We set $r = 2$ and target $k = 20$ nonzero loadings per component. To avoid
time-consuming parameter tuning, we consider only methods that accept exact cardinality
budgets.

| Method | $(k_1, k_2)$ | FVE | Orth. violation | Runtime (s) | Memory (MB) |
|---|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (20,20) | 0.008 | < 1e-15 | 85.713 | 640 [640, 641] |
| `msPCA::mspca` $(\boldsymbol{X})$ | (20,20) | 0.008 | < 1e-15 | 2.167 | 44 [42, 49] |
| `elasticnet::spca` $(\boldsymbol{X})$ | (20,20) | 0.004 | < 1e-15 | 6.277 | 476 [309, 645] |
| `mixOmics::spca` | (20,20) | 0.006 | < 1e-15 | 0.345 | 24 [17, 50] |
| `nsprcomp::nsprcomp` | (20,20) | 0.009 | < 1e-15 | 0.637 | 108 [54, 198] |
| PCA (dense) | (4088,4088) | 0.478 | < 1e-15 | 0.080 | 42 [39, 43] |

`msPCA` is run with `maxIter = 100` here; all methods other than `msPCA` $(\boldsymbol{\Sigma})$ use the
data matrix directly. Only `msPCA` varies at all across seeds, and only in the third decimal of
FVE (0.008--0.009 in both input modes).

This dataset isolates the benefit of the `type = "X"` interface. Passing the raw data matrix
replaces the $O(p^2)$ product $\boldsymbol{\Sigma}\boldsymbol{\beta}$ with the $O(np)$ two-pass product
$\boldsymbol{X}^\top(\boldsymbol{X}\boldsymbol{\beta})/(n-1)$ and never forms the covariance matrix. The measured working
set falls from 640 MB to 44 MB, a factor of about 15, at identical FVE and feasibility. The
$p \times p$ correlation matrix accounts for 128 MB of that; the remainder is the `Eigen` copy of
it and the dense intermediates the $O(p^2)$ path requires. Runtime falls from 85.7 seconds to
2.2, a factor of 40. Both `msPCA` rows are also the most stable in the table --- 640 [640, 641]
and 44 [42, 49] --- while `elasticnet` spans 309--645 MB and `nsprcomp` 54--198 MB across
repetitions of identical work.

That stability is not incidental. `msPCA` allocates its working memory in Eigen, where the
footprint is determined by the algorithm; the pure-R methods allocate on R's heap, where the
peak depends on when the garbage collector happens to run. The practical consequence is that
`msPCA`'s memory requirement can be predicted from $n$, $p$ and $r$, and the others' cannot.

Read together with the Pitprops result, the rule is straightforward: use `type = "X"` when
$n \ll p$ and `type = "Sigma"` when $p \lesssim n$, which is what the relative sizes of
$\boldsymbol{X}$ and $\boldsymbol{\Sigma}$ would suggest.

## Discussion

**Sparsity control.** `PMA::SPC()`, `sparsepca::spca()`, `nsprcomp::nscumcomp()`,
`amanpg::spca.amanpg()` and `sklearn.decomposition.SparsePCA()` do not provide satisfactory
control over the sparsity of each loading vector, which makes an apples-to-apples comparison
difficult. `PMA::SPC()` and `sklearn` control only a global $\ell_1$ bound, while
`sparsepca::spca()`, `nscumcomp()` and `amanpg` tend to produce unbalanced components with one
PC carrying almost all of the budget. It is difficult to calibrate these methods to an exact
target number of nonzeros per component. The comparisons below therefore focus on
`msPCA::mspca()`, `elasticnet::spca()`, `mixOmics::spca()` and `nsprcomp::nsprcomp()`.

**Feasibility.** Among the packages considered, `msPCA::mspca()` is the only method that
consistently returns small orthogonality violations on all four datasets, at or below 1e-4
(its default feasibility tolerance). Because every method is now run at several seeds, this is
a worst-case statement rather than a single-run one: across all repetitions on all four
datasets, the largest violation `msPCA` returned was 9.9e-05. All other packages can return
substantially non-orthogonal components, especially when $p$ is small relative to the total
sparsity budget, so that trivially orthogonal disjoint-support solutions may not exist.

**Variance explained.** `msPCA::mspca()` and `nsprcomp::nsprcomp()` are generally the
best-performing methods, followed by `mixOmics::spca()` and `elasticnet::spca()`, though the
relative ranking of the latter two varies across datasets. `msPCA` does not appear to sacrifice
variance explained in order to achieve feasibility. The one method that outperforms it on FVE
while respecting the cardinality budget is `nscumcomp` on Pitprops, which reaches 0.845 against
0.791 --- but with an orthogonality violation of 1.9, and with a spread of 0.750--0.905 across
seeds, so the comparison is not like for like.

**Reproducibility.** Repeating each method under several seeds separates the deterministic
implementations (`elasticnet`, `PMA`, `sparsepca`, `mixOmics`, `amanpg`, dense PCA) from those
that are not. `nsprcomp` and `nscumcomp` use random initialisation and can move substantially
between runs --- `nscumcomp`'s FVE spans 0.750--0.905 on Pitprops and its orthogonality
violation 1.9--2.7. `msPCA` uses random restarts and is also not bit-reproducible across seeds,
but its variation is confined to the objective: the feasibility guarantee holds at every seed on
every dataset, the largest violation observed anywhere being 9.9e-05.

**Memory.** On the two high-dimensional datasets, where the memory column reflects the
algorithms rather than fixed overheads, `msPCA` is the lightest of the sparse methods that
respect an exact cardinality budget. On breast cancer it uses 3.0 MB on the data matrix against
174 MB for `elasticnet` on the same input and 282 MB in Gram mode, with `sparsepca`, `nsprcomp`,
`nscumcomp` and `amanpg` all in the 48--56 MB range. On riboflavin it uses 44 MB against 476 MB
for `elasticnet` and 108 MB for `nsprcomp`. The $\boldsymbol{\Sigma}$-versus-$\boldsymbol{X}$ choice matters more
than the choice of package: `msPCA` itself spans 640 MB to 44 MB on riboflavin depending on which
interface is used.

`msPCA`'s memory figures are also markedly more stable, because it allocates in Eigen rather
than on R's heap, where peak footprint depends on garbage-collector timing. Across five
repetitions of identical work on riboflavin, `msPCA` spans 640--641 MB and 42--49 MB in its two
modes, while `elasticnet` spans 309--645 MB and `nsprcomp` 54--198 MB. For a user sizing a job,
a predictable requirement is worth as much as a small one.

**Scalability.** `msPCA::mspca()` scales to datasets with $p$ in the thousands in reasonable
time, and the `type = "X"` interface yields substantial runtime and memory gains when
$n \ll p$ --- 40x faster and 15x lighter on riboflavin. When $n > p$, as in the Pitprops
pseudo-data, the ordering reverses and `type = "Sigma"` is preferable.

## References
