---
title: "Benchmarking against other sparse PCA packages"
output: rmarkdown::html_vignette
bibliography: thebib.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette gives the full benchmarking protocol and results summarised in the package
description. It compares `msPCA` against seven competing implementations on four real
datasets, and documents exactly how each competitor was configured --- which matters, because
the packages parameterise sparsity in incompatible ways.

Chunks are not evaluated at build time: reproducing the full comparison requires eight
additional packages plus a Python installation. The complete scripts live in the package
repository under
[`replication/benchmarking/`](https://github.com/jeanpauphilet/msPCA/tree/main/replication/benchmarking).

## Getting the data

All four datasets are obtainable from R without manual downloads or accounts, so the
comparison is reproducible end to end:

```{r}
## mtcars -- base R
X_mtcars     <- as.matrix(datasets::mtcars)
Sigma_mtcars <- cor(X_mtcars)

## Pitprops -- the 13 x 13 correlation matrix of Jeffers (1967), Table 1.
## It is transcribed literally at the top of
## replication/benchmarking/notebook_pitprops.R; source that file to obtain `pitprops`.
Sigma_pitprops <- pitprops
## Methods that require a data matrix instead use pseudo-data drawn from this
## correlation structure (see "How each method was configured" below).
set.seed(42)
X_pitprops <- MASS::mvrnorm(n = 500, mu = rep(0, ncol(pitprops)),
                            Sigma = pitprops)

## Breast cancer -- one-off download via PMA, then keep the 500 highest-
## variance genes, following the preprocessing of Witten et al. (2009).
X            <- t(PMA::download_breast_data(
  url = "https://tibshirani.su.domains/PMA/breastdata.rda")$rna)
X_breast     <- X[, order(apply(X, 2, var), decreasing = TRUE)[1:500]]
Sigma_breast <- cor(X_breast)          # n = 89, p = 500

## Riboflavin -- ships with the hdi package
data("riboflavin", package = "hdi")
X_ribo <- as.matrix(riboflavin$x)      # n = 71, p = 4088
```

Only the breast cancer data touches the network, and `PMA::download_breast_data()` handles it
in a single call. Riboflavin ships inside `hdi`, `mtcars` is in base R, and the Pitprops
correlation matrix is transcribed in the benchmarking script, so no manual downloads or
accounts are needed anywhere.

## Methods compared

We compare `msPCA` against seven competing methods:

| Package | Reference | Language | Approach |
|---|---|---|---|
| `elasticnet` | @zou2006sparse | R | elastic-net regularization, LARS-like |
| `PMA` | @witten2009penalized | R | penalized matrix decomposition |
| `sparsepca` | @erichson2020sparse | R | variable projection with $\ell_1$ penalty |
| `mixOmics` | @rohart2017mixomics | R | regularized SVD [@shen2008sparse] |
| `nsprcomp` | @sigg2019nsprcomp | R | expectation-maximization [@sigg2008expectation] + deflation |
| `amanpg` | @chen2020proximal | R | alternating manifold proximal gradient |
| `sklearn.decomposition.SparsePCA` | @pedregosa2011scikit | Python | dictionary learning |

`scikit-learn` is called from R via the `reticulate` package [@ushey2023reticulate]. Together
these cover the main methodological approaches to sparse PCA: elastic-net regularization,
penalized matrix decomposition, variable-projection optimization, regularized SVD, deflation,
Riemannian proximal gradient, and dictionary learning.

## Environment

Experiments were conducted on a MacBook Air with an Apple M2 chip and 24 GB of RAM, with
R version 4.6.0 (2026-04-24), `msPCA` v0.5.0, `elasticnet` v1.3, `PMA` v1.2.4, `sparsepca`
v0.1.2, `mixOmics` v6.36.0, `nsprcomp` v0.5.1.2, `amanpg` v0.1.0, `reticulate` v1.40.0, and
`scikit-learn` v1.6.1 (Python 3.11). Full session information is recorded in
[`replication/sessionInfo.txt`](https://github.com/jeanpauphilet/msPCA/blob/main/replication/sessionInfo.txt).

## How each method was configured

Comparing sparse PCA methods is complicated because different packages parameterize sparsity
differently. Some accept a covariance or correlation matrix, others apply only to the data
matrix, and some support both.

- **`msPCA`** --- `mspca()` with both `type = "Sigma"` and `type = "X"` on all four datasets.
  On riboflavin ($n \ll p$) the $p \times p$ covariance matrix is rank-deficient, making
  `type = "X"` the natural choice; we report both modes to quantify the runtime gain.
- **`elasticnet`** --- `spca()`, which takes a list of cardinality budgets (one per PC) and
  applies interchangeably to $\boldsymbol{\Sigma}$ or $\boldsymbol{X}$. We report both input modes on mtcars,
  Pitprops and breast cancer.
- **`PMA`** --- `SPC()` with `orth = TRUE`. It takes the covariance matrix and a bound
  `sumabsv` on $\sum_t \|\boldsymbol{u}_t\|_1$, which we tune manually per dataset to get
  per-component sparsity as close as possible to the target.
- **`sparsepca`** --- `spca()` applied to $\boldsymbol{\Sigma}$, which only exposes the magnitude of
  the $\ell_1$ penalty `alpha`. We tune it per dataset to achieve a total sparsity
  $\sum_t k_t$ close to the target. This method tends to return PCs with very unequal sparsity
  levels, the first typically denser than the rest.
- **`mixOmics`** --- `spca()`, which takes the data matrix and the list of sparsity budgets.
- **`nsprcomp`** --- two functions, both based on the EM algorithm of @sigg2008expectation.
  `nsprcomp()` computes the leading sparse PC then applies the deflation procedure of
  @mackey2008deflation for subsequent PCs; it takes the data matrix and a list of sparsity
  levels. `nscumcomp()` solves for all $r$ PCs simultaneously, taking a total sparsity budget
  $\sum_t k_t$ and a penalty parameter `gamma` on orthogonality violation (default 0).
- **`amanpg`** --- `spca.amanpg()` applied to the correlation matrix (`type = 1`). It solves a
  single-component problem with an $\ell_1$ penalty `lambda1` and applies sequential deflation;
  we tune `lambda1` per dataset.
- **`scikit-learn`** --- `sklearn.decomposition.SparsePCA()` via `reticulate`. The class
  computes $r$ components simultaneously using dictionary learning with an $\ell_1$ penalty
  `alpha` on the loadings, applied to the data matrix. We tune `alpha` per dataset.

For Pitprops only the correlation matrix is available, so we generate pseudo-data via
`MASS::mvrnorm()` with the Pitprops matrix as the population covariance ($n = 500$) for the
methods that require $\boldsymbol{X}$.

For each method we report the actual number of nonzero loadings per PC, the FVE computed on
the original correlation matrix, and the orthogonality violation. Runtimes exclude
hyperparameter tuning and are measured with `system.time()`. Metrics come from a single run
per method and dataset, with all seeds fixed in the replication scripts.

```{r}
# Representative call pattern; see replication/benchmarking/ for the full scripts.
set.seed(42)
res <- mspca(Sigma, r = 3, ks = rep(4, 3), verbose = FALSE)

fraction_variance_explained(Sigma, res$x_best)
feasibility_violation_off(Sigma, res$x_best, 0)
res$runtime
```

## Results

### mtcars

`datasets::mtcars` from base R ($p = 11$ variables, $n = 32$ observations) serves as a simple
reproducible baseline. We extract $r = 3$ sparse PCs targeting $k = 4$ per component.

| Method | $(k_1,k_2,k_3)$ | FVE | Orth. violation | Runtime (s) |
|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (4,4,4) | 0.829 | 9.4e-05 | 0.323 |
| `msPCA::mspca` $(\boldsymbol{X})$ | (4,4,4) | 0.829 | 9.4e-05 | 0.424 |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (4,4,4) | 0.753 | 0.117 | 0.165 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (4,4,4) | 0.754 | 0.117 | 0.189 |
| `PMA::SPC` | (4,4,5) | 0.552 | 0.283 | 0.008 |
| `sparsepca::spca` | (7,4,1) | 0.585 | 0.045 | 0.053 |
| `mixOmics::spca` | (4,4,4) | 0.633 | 0.133 | 0.003 |
| `nsprcomp::nsprcomp` | (4,4,4) | 0.832 | 0.078 | 0.006 |
| `nsprcomp::nscumcomp` | (6,3,3) | 0.772 | 0.245 | 0.033 |
| `amanpg::spca.amanpg` | (7,4,2) | 0.649 | 0.041 | 0.004 |
| `sklearn...SparsePCA` | (4,4,3) | 0.827 | < 1e-15 | 0.028 |
| PCA (dense) | (11,11,11) | 0.899 | < 1e-15 | --- |

Tuning: `PMA` uses `sumabsv = 1.69`; `sparsepca` uses `alpha = 0.004`; `amanpg` uses
`lambda1 = c(10, 3, 0.01)`; `scikit-learn` uses `alpha = 3.2`.

### Pitprops

The Pitprops dataset [@jeffers1967two] ($p = 13$ physical measurements on $n = 180$ timber
specimens) is a classic sparse PCA benchmark [e.g. @zou2006sparse]. We extract $r = 6$ sparse
PCs targeting $k = 4$ nonzero loadings per component, as in @zou2006sparse.

| Method | $(k_1, \ldots, k_6)$ | FVE | Orth. violation | Runtime (s) |
|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (4,4,4,4,4,4) | 0.786 | 9.7e-05 | 1.109 |
| `msPCA::mspca` $(\boldsymbol{X})$ | (4,4,4,4,4,4) | 0.801 | 9.5e-05 | 5.388 |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (4,4,4,4,4,4) | 0.792 | 0.241 | 0.070 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (4,4,4,4,4,4) | 0.790 | 0.322 | 0.167 |
| `PMA::SPC` | (4,4,4,7,6,4) | 0.660 | 0.857 | 0.011 |
| `sparsepca::spca` | (8,4,3,1,3,5) | 0.762 | 0.376 | 0.024 |
| `mixOmics::spca` | (4,4,4,4,4,4) | 0.788 | 0.215 | 0.008 |
| `nsprcomp::nsprcomp` | (4,4,4,4,4,4) | 0.800 | 0.626 | 0.023 |
| `nsprcomp::nscumcomp` | (5,2,4,5,3,5) | 0.737 | 1.691 | 1.072 |
| `amanpg::spca.amanpg` | (8,6,9,2,3,5) | 0.760 | 1.763 | 0.008 |
| `sklearn...SparsePCA` | (7,3,4,2,4,4) | 0.895 | 0.733 | 0.043 |
| PCA (dense) | (13,...,13) | 0.877 | < 1e-15 | --- |

Tuning: `PMA` uses `sumabsv = 1.727`; `sparsepca` uses `alpha = 0.004`; `nscumcomp` uses the
default `gamma = 0`; `amanpg` uses `lambda1 = c(1, 1, 5, 0.4, 0.4, 0)`; `scikit-learn` uses
`alpha = 3.1`.

This is the dataset where the feasibility differences are starkest: competitors return
orthogonality violations in the 0.2--1.8 range, against 9.7e-05 for `msPCA`. With $p = 13$ and
a total budget of 24 nonzeros, trivially orthogonal disjoint-support solutions do not exist,
so methods that rely on deflation have nothing to fall back on.

### Breast cancer gene expression

Gene expression data from @chin2006genomic ($n = 89$ tumor samples, 19,672 genes). Following
the preprocessing of @witten2009penalized we retain the $p = 500$ genes with the highest
marginal variance, yielding a high-dimensional ($p > n$) benchmark. We set $r = 3$ and target
$k = 20$ nonzero loadings per component.

| Method | $(k_1, k_2, k_3)$ | FVE | Orth. violation | Runtime (s) |
|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (20,20,20) | 0.093 | 7.5e-05 | 27.472 |
| `msPCA::mspca` $(\boldsymbol{X})$ | (20,20,20) | 0.093 | 7.5e-05 | 13.382 |
| `elasticnet::spca` $(\boldsymbol{\Sigma})$ | (20,20,20) | 0.043 | 0.040 | 18.130 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (20,20,20) | 0.043 | 0.021 | 0.904 |
| `PMA::SPC` | (14,17,27) | 0.028 | 0.001 | 0.042 |
| `sparsepca::spca` | (46,13,8) | 0.061 | < 1e-15 | 4.752 |
| `mixOmics::spca` | (20,20,20) | 0.077 | < 1e-15 | 0.031 |
| `nsprcomp::nsprcomp` | (20,20,20) | 0.084 | < 1e-15 | 0.027 |
| `nsprcomp::nscumcomp` | (33,22,5) | 0.082 | < 1e-15 | 0.152 |
| `amanpg::spca.amanpg` | (96,4,2) | 0.067 | 0.063 | 0.252 |
| `sklearn...SparsePCA` | (56,3,6) | 0.057 | < 1e-15 | 0.180 |
| PCA (dense) | (500,500,500) | 0.352 | < 1e-15 | --- |

Tuning: `PMA` uses `sumabsv = 2.92`; `sparsepca` uses `alpha = 0.004` but returns unequal
sparsity levels; `nscumcomp` receives a total budget of $k = 60$ and distributes it unevenly;
`amanpg` uses `lambda1 = c(8.5, 2.5, 1.5)`; `scikit-learn` uses `alpha = 11.25`.

With $p = 500$ and only 60 nonzeros in total, disjoint supports are easy to find, so most
methods are trivially orthogonal here.

### Riboflavin

The riboflavin (vitamin B2) production dataset [@buhlmann2014high] comprises log-transformed
expression levels of $p = 4{,}088$ genes in $n = 71$ samples of *Bacillus subtilis*. This is a
challenging $p \gg n$ benchmark: the empirical correlation matrix is rank-deficient
(rank $\leq 70$). We set $r = 2$ and target $k = 20$ nonzero loadings per component. To avoid
time-consuming parameter tuning, we consider only methods that accept exact cardinality
budgets.

| Method | $(k_1, k_2)$ | FVE | Orth. violation | Runtime (s) |
|---|---|---|---|---|
| `msPCA::mspca` $(\boldsymbol{\Sigma})$ | (20,20) | 0.008 | < 1e-15 | 71.986 |
| `msPCA::mspca` $(\boldsymbol{X})$ | (20,20) | 0.008 | < 1e-15 | 1.603 |
| `elasticnet::spca` $(\boldsymbol{X})$ | (20,20) | 0.004 | < 1e-15 | 5.952 |
| `mixOmics::spca` | (20,20) | 0.006 | < 1e-15 | 0.241 |
| `nsprcomp::nsprcomp` | (20,20) | 0.009 | < 1e-15 | 0.351 |
| PCA (dense) | (4088,4088) | 0.478 | < 1e-15 | 0.045 |

`msPCA` is run with `maxIter = 100` here; all methods other than `msPCA` use the data matrix
directly.

This dataset isolates the benefit of the `type = "X"` interface. Passing the raw data matrix
replaces the $O(p^2)$ product $\boldsymbol{\Sigma}\boldsymbol{\beta}$ with the $O(np)$ two-pass product
$\boldsymbol{X}^\top(\boldsymbol{X}\boldsymbol{\beta})/(n-1)$ and avoids forming the covariance matrix at all --- for
$p = 4{,}088$ that is roughly 134 MB of double-precision storage never allocated. Runtime falls
from about 72 seconds to under 2, making `msPCA` faster than `elasticnet::spca()` on this
problem.

## Discussion

**Sparsity control.** `PMA::SPC()`, `sparsepca::spca()`, `nsprcomp::nscumcomp()`,
`amanpg::spca.amanpg()` and `sklearn.decomposition.SparsePCA()` do not provide satisfactory
control over the sparsity of each loading vector, which makes an apples-to-apples comparison
difficult. `PMA::SPC()` and `sklearn` control only a global $\ell_1$ bound, while
`sparsepca::spca()`, `nscumcomp()` and `amanpg` tend to produce unbalanced components with one
PC carrying almost all of the budget. It is difficult to calibrate these methods to an exact
target number of nonzeros per component. The comparisons below therefore focus on
`msPCA::mspca()`, `elasticnet::spca()`, `mixOmics::spca()` and `nsprcomp::nsprcomp()`.

**Feasibility.** Among the packages considered, `msPCA::mspca()` is the only method that
consistently returns small orthogonality violations on all four datasets, at or below 1e-4
(its default feasibility tolerance). All other packages can return substantially
non-orthogonal components, especially when $p$ is small relative to the total sparsity budget,
so that trivially orthogonal disjoint-support solutions may not exist.

**Variance explained.** `msPCA::mspca()` and `nsprcomp::nsprcomp()` are generally the
best-performing methods, followed by `mixOmics::spca()` and `elasticnet::spca()`, though the
relative ranking of the latter two varies across datasets. `msPCA` does not appear to sacrifice
variance explained in order to achieve feasibility.

**Scalability.** `msPCA::mspca()` scales to datasets with $p$ in the thousands in reasonable
time, and the `type = "X"` interface yields substantial runtime and memory gains when
$n \ll p$.

## References
